---
title: "Statistical thinking for wet-lab experts"
author: "Jitao David Zhang"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    code_folding: hide
    fig_width: 7
    fig_height: 4.5
params:
  seed: 20260919
  n_sim: 2000          # Monte Carlo replicates for modules 6 and 8
  outdir: "output/workshop"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "png", dpi = 200,
                      ## figures are drawn at their final size: 7 inches or
                      ## less, so 12 pt text really is 12 pt on the page
                      fig.width = 7, fig.height = 4.5,
                      ## keep every figure on disk: the slide deck and the
                      ## one-pager are built from these files by the Makefile
                      fig.path = "output/figures/")
```

```{r libraries}
library(tidyverse)
library(patchwork)
library(ribiosUtils)   # ofactor, percentage
library(ribiosPlot)    # openFileDevice, closeFileDevice
library(gt)

source("R/simulate_workshop_data.R")
set.seed(params$seed)
dir.create(params$outdir, showWarnings = FALSE, recursive = TRUE)

## One theme for the whole document. Nothing is below 12 pt, and every figure
## is drawn at 7 inches wide or less, so the text keeps a generous share of
## the canvas instead of shrinking when the figure is placed on a slide.
BASE_PT <- 12                  # smallest type anywhere in a figure
GEOM_PT <- BASE_PT / .pt       # the same size for geom_text()/annotate(), mm

theme_ws <- theme_bw(base_size = BASE_PT) +
  theme(plot.title    = element_text(face = "bold", size = BASE_PT + 1.5,
                                     margin = margin(b = 1)),
        plot.subtitle = element_text(size = BASE_PT, colour = "grey30",
                                     margin = margin(b = 3)),
        axis.title.x  = element_text(size = BASE_PT, margin = margin(t = 3)),
        axis.title.y  = element_text(size = BASE_PT, margin = margin(r = 3)),
        axis.text     = element_text(size = BASE_PT),
        strip.text    = element_text(size = BASE_PT, face = "bold",
                                     margin = margin(2, 2, 2, 2)),
        legend.text   = element_text(size = BASE_PT),
        legend.title  = element_blank(),
        plot.title.position = "plot",
        ## the defaults leave a lot of air around the legend and between
        ## facets; most of the empty space in a small figure lives here
        legend.key.size    = unit(0.8, "lines"),
        legend.margin      = margin(0, 0, 0, 0),
        legend.box.spacing = unit(3, "pt"),
        panel.spacing      = unit(4, "pt"),
        plot.margin        = margin(3, 5, 3, 3),
        panel.grid.minor = element_blank())

## Discrete axes default to half a category of padding at each end; 0.45
## keeps the outermost box clear of the panel edge without the wide gutters.
## These are functions, not stored scale objects: a ggplot2 scale is mutable,
## so adding one object to several plots would let the axis levels of one
## leak into the next.
tight_x <- function(add = 0.45) scale_x_discrete(expand = expansion(add = add))
tight_y <- function(add = 0.45) scale_y_discrete(expand = expansion(add = add))
flush   <- expansion(add = 0)          # for tile plots, which should fill
theme_set(theme_ws)

cpd_cols <- c("CPD-A" = "#1f6fb4", "CPD-B" = "#c1443b")
day_cols <- c("Friday" = "#e07b39", "Monday" = "#4c8fbd")
ok_cols  <- c("yes" = "#3b7a57", "no" = "#a33", "maybe" = "#d9a441")

## One picture, used in five modules: an estimate, the range of differences the
## data are compatible with, and the line we decided to act on. Reusing the
## same figure is the point - reading an interval against a threshold is the
## habit this workshop is trying to build.
verdict_cols <- c("relevant difference" = "#3b7a57",
                  "no relevant difference" = "#4c8fbd",
                  "inconclusive" = "#d9a441")

forest <- function(d, title, subtitle = NULL, threshold = MCID_FOLD,
                   xlab = "CPD-B / CPD-A  (cytotoxicity IC50 fold change)") {
  d <- d %>%
    mutate(row = factor(label, levels = rev(label)),
           verdict = case_when(is.na(lwr)      ~ "inconclusive",
                               lwr > threshold ~ "relevant difference",
                               upr < threshold ~ "no relevant difference",
                               TRUE            ~ "inconclusive"),
           verdict = factor(verdict, levels = names(verdict_cols)))
  ggplot(d, aes(estimate, row, colour = verdict)) +
    ## xmin = 0 (not -Inf): after the log transform this is the panel edge,
    ## whereas -Inf becomes NaN and the band silently disappears
    annotate("rect", xmin = 0, xmax = threshold, ymin = -Inf, ymax = Inf,
             fill = "grey93") +
    geom_vline(xintercept = threshold, linetype = "dashed", linewidth = 0.9) +
    geom_vline(xintercept = 1, colour = "grey70") +
    geom_linerange(aes(xmin = lwr, xmax = upr), linewidth = 1.3, na.rm = TRUE) +
    geom_point(size = 2.8, na.rm = TRUE) +
    scale_x_log10() +
    scale_colour_manual(values = verdict_cols, drop = FALSE) +
    scale_y_discrete(expand = expansion(add = 0.45)) +
    labs(title = title, subtitle = subtitle, x = xlab, y = NULL) +
    theme(legend.position = "bottom")
}

## every module exports the data it uses, so participants can reopen it
## afterwards in Excel, Prism, or with an LLM
.manifest <- tibble(file = character(), rows = integer(), what = character())

## keep exported numbers readable without corrupting raw counts or p-values
tidy_num <- function(x) {
  if (!is.double(x)) return(x)
  nz <- abs(x[is.finite(x) & x != 0])
  if (length(nz) && min(nz) < 1e-3) signif(x, 3) else round(x, 3)
}

export_data <- function(x, name, what) {
  path <- file.path(params$outdir, paste0(name, ".csv"))
  write_csv(mutate(x, across(where(is.double), tidy_num)), path)
  .manifest <<- add_row(.manifest, file = basename(path),
                        rows = nrow(x), what = what)
  invisible(x)
}
```

```{r the-data}
sim   <- simulate_workshop_data(seed = params$seed)
viab  <- sim$viab                 # 48 wells, the dataset the workshop uses
dm    <- lot_summary(viab)        # 16 lot-level summaries, the analysis unit
folds <- lot_fold(viab)           # 8 within-lot fold shifts, the estimand
thr   <- sim$thresholds
CONC  <- sim$params$screen_conc
HILL  <- sim$params$hill_A

## Two scales, and it matters which is which.
##   MCID_FOLD  the decision threshold: a 3-fold shift in cytotoxicity IC50.
##              This is the scale the project actually decides on, and the
##              scale on which the effect is additive, so every comparison
##              from Module III onwards is made here.
##   MCID_PP    the same threshold expressed in percentage points of
##              viability, for Modules I and II where we are still talking
##              about what the instrument reports.
MCID_FOLD <- thr$mcid_fold
MCID_PP   <- thr$mcid_pp

## a paired comparison of the two compounds, on the log scale where the lot
## effect cancels, reported back as a fold
fold_test <- function(d, conf = 0.95) {
  tt <- t.test(log10(d$`CPD-B`), log10(d$`CPD-A`), paired = TRUE, conf.level = conf)
  tibble(estimate = 10^unname(tt$estimate),
         lwr = 10^tt$conf.int[1], upr = 10^tt$conf.int[2],
         p_value = tt$p.value)
}
```

# How to use this notebook

Every module follows the same five-step loop, and the loop is the point of
the workshop:

> **Question → Design → Data → Analysis and uncertainty → Decision**

One dataset carries all nine modules. It is synthetic, so the true answer is
known and every claim in here can be checked against it.

**The experiment.** Two compounds from the same project, CPD-A and CPD-B, are
tested for cytotoxicity in primary human hepatocytes with the CellTiter-Glo
ATP readout.

| | |
|---|---|
| Independent units | 8 hepatocyte lots (donors) |
| Conditions | CPD-A and CPD-B, both in every lot |
| Technical replicates | 3 wells per lot × compound |
| Readout concentration | `r CONC` µM, one point on the curve |
| Run days | Friday and Monday |
| Total measurements | 48 |

Three things are hidden in it, and each surfaces in the module where it
matters: a run-day effect that lives in the vehicle reference, one well with
an air bubble, and one lot that is genuinely more sensitive than the rest.

> **A note on the numbers.** This notebook is the participants' document. A
> companion document, `2026-09-statistical-thinking.Rmd`, works the same assay
> at full analyst depth: 96-well plates, eight concentrations, four-parameter
> logistic fits, plate QC, weighted regression and variance decomposition.
> Modules IV and VI point into it. The two documents use the same compounds and
> the same day effect but are simulated separately, so the individual numbers
> differ.

---

# Module I — What questions are we asking, and what is the minimal difference that we care about?

### Story

A project team wants to know whether they should continue developing the candidate CPD-B, depending on the question whether CPD-B increases IC50 in the cytotoxity data at at least three folds. If someone says "run the viability assay and tell me if it's significant", we cannot tell: because it does not say what would count as significant.

Five questions turn it into an experiment:

* Measurements **of what**? — ATP content, as a proxy for viable cells
* From **whom or what**? — primary human hepatocytes, one lot per donor
* Under **which conditions**? — `r CONC` µM, 24 h, 0.1% DMSO vehicle
* What **population** do we want to speak about? — human hepatocyte lots in
  general, not these eight
* What **decision** does this inform? — whether CPD-B needs a dedicated
  safety study

### The sentence every experiment should be able to complete

> I want to compare **viability at `r CONC` µM** in **primary human
> hepatocyte lots** because I need to decide **whether CPD-B is substantially
> less cytotoxic than CPD-A**.

### The threshold comes before the data

The team agrees, before seeing anything, that a **`r thr$mcid_fold`-fold shift
in the cytotoxicity IC~50~** is what they would act on. That is a scientific
judgement, not a statistical one — no analysis can supply it.

A word on the term. This assay says nothing about a target: it reports how much
compound is needed to kill cells. So the quantity is the **IC~50~ for
cytotoxicity**, not potency — potency is normally defined against a target, and
using it here would claim something the experiment cannot support.

Two scales are now in play, and the difference between them runs through the
whole workshop:

* **what the instrument reports** — viability at one concentration, in
  percentage points;
* **what the project decides on** — the concentration that kills half the
  cells, in µM, compared as a fold shift.

The second is a property of the compound; the first is a property of the
compound *and* the concentration we happened to choose. Converting between
them is just algebra on the Hill equation, but it needs the slope *h*:

$$\mathrm{IC}_{50} = c \left(\frac{v}{100 - v}\right)^{1/h}$$

and it is steep at the ends — a lot sitting near 0% or 100% viability gives an
IC~50~ that moves a great deal for a small change in the reading.

```{r m1-threshold-prepare}
## --- A: the two compounds, and the shift read horizontally --------------
curve_ab <- expand_grid(
    tibble(compound = c("CPD-A", "CPD-B"),
           ic50     = c(sim$params$ic50_A, sim$params$ic50_B)),
    conc = 10^seq(log10(0.3), log10(1000), length.out = 300)) %>%
  mutate(viability = 100 * hill_viability(conc, ic50, sim$params$hill_A))

ic50_A <- sim$params$ic50_A
ic50_B <- sim$params$ic50_B
ic50_T <- ic50_A * thr$mcid_fold          # where the 3-fold threshold sits

p1a <- ggplot(curve_ab, aes(conc, viability, colour = compound)) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 50, linetype = "dotted", colour = "grey55") +
  ## the shift as a pharmacologist reads it: along the 50% line, left to right
  annotate("segment", x = ic50_A, xend = ic50_B, y = 50, yend = 50,
           linewidth = 1.1, colour = "grey25",
           arrow = arrow(ends = "both", length = unit(0.16, "cm"))) +
  annotate("label", x = sqrt(ic50_A * ic50_B), y = 50,
           label = sprintf("%.0f-fold", thr$true_fold), size = GEOM_PT,
           label.padding = unit(0.02, "lines")) +
  ## and where the decision threshold would have been
  annotate("segment", x = ic50_T, xend = ic50_T, y = 0, yend = 50,
           linetype = "dashed", colour = "#7a3b8f", linewidth = 0.9) +
  annotate("text", x = ic50_T, y = 14, hjust = -0.08, vjust = 0,
           label = "3-fold", size = GEOM_PT, colour = "#7a3b8f") +
  scale_x_log10() +
  scale_colour_manual(values = cpd_cols) +
  labs(title = "Q: is there a substantial shift in IC50?",
       x = "Concentration (µM)", y = "ATP (% of vehicle)") +
  theme(legend.position = "bottom")

## --- B: the same shift, read vertically at one concentration ------------
series_levels <- c("CPD-A", "3\u00d7 higher IC50")
curve_df <- expand_grid(
    series = series_levels,
    conc   = 10^seq(log10(0.3), log10(1000), length.out = 300)) %>%
  mutate(ic50 = if_else(series == "CPD-A",
                        sim$params$ic50_A,
                        sim$params$ic50_A * thr$mcid_fold),
         viability = 100 * hill_viability(conc, ic50, sim$params$hill_A)) %>%
  mutate(series = factor(series, levels=series_levels))

v_lo <- 100 * hill_viability(CONC, sim$params$ic50_A, sim$params$hill_A)
v_hi <- 100 * hill_viability(CONC, sim$params$ic50_A * thr$mcid_fold,
                             sim$params$hill_A)

p1a_prime <- ggplot(curve_df, aes(conc, viability, colour = series, linetype = series)) +
  geom_line(linewidth = 1.3) +
  geom_hline(yintercept = 50, linetype = "dotted", colour = "grey55") +
  ## the shift as a pharmacologist reads it: along the 50% line, left to right
  annotate("segment", x = ic50_A, xend = ic50_A * thr$mcid_fold, y = 50, yend = 50,
           linewidth = 1.1, colour = "grey25",
           arrow = arrow(ends = "both", length = unit(0.16, "cm"))) +
  ## and where the decision threshold would have been
  annotate("text", x = ic50_T, y = 50, hjust = -0.08, vjust = 0,
           label = "3-fold", size = GEOM_PT, colour = "grey25") +
  scale_x_log10(breaks=c(1, 10, 30, 100, 1000)) +
  scale_colour_manual(values = c("CPD-A" = "#1f6fb4",
                                 "3\u00d7 higher IC50" = "#7a3b8f")) +
  scale_linetype_manual(values = c("CPD-A"=1, "3\u00d7 higher IC50"=2)) +
  labs(title = "Q: is there a substantial shift in IC50?",
       x = "Concentration (µM)", y = "ATP (% of vehicle)") +
  theme(legend.position = "bottom")

p1b <- ggplot(curve_df, aes(conc, viability, colour = series, linetype=series)) +
  geom_line(linewidth = 1.3) +
  annotate("segment", x = CONC, xend = CONC, y = v_lo, yend = v_hi,
           linewidth = 1.2, colour = "grey25",
           arrow = arrow(ends = "both", length = unit(0.18, "cm"))) +
  annotate("label", x = CONC, y = (v_lo + v_hi) / 2,
           label = sprintf("%.0f pp", v_hi - v_lo), size = GEOM_PT) +
  geom_vline(xintercept = CONC, linetype = "dotted", colour = "grey40") +
  scale_x_log10() +
  scale_colour_manual(values = c("CPD-A" = "#1f6fb4",
                                 "3\u00d7 higher IC50" = "#7a3b8f")) +
  scale_linetype_manual(values = c("CPD-A"=1, "3\u00d7 higher IC50"=2)) +

  labs(title = "Simplified to single-conc. measurement",
       subtitle = sprintf("38 percentage points (pp) is equivalent to %dx IC50",
                          thr$mcid_fold),
       x = "Concentration (µM)", y = "ATP (% of vehicle)") +
  theme(legend.position = "bottom")

p1c <- sim$mcid_by_donor %>%
  ggplot(aes(fct_reorder(donor_id, mcid_pp), mcid_pp)) +
  geom_col(fill = "#7a3b8f", width = 0.82) +
  scale_x_discrete(expand = expansion(add = 0.55)) +
  geom_hline(yintercept = MCID_PP, linetype = "dashed", linewidth = 1) +
  geom_hline(yintercept = thr$mcid_pp_naive, linetype = "dotted",
             colour = "grey35", linewidth = 1) +
  annotate("text", x = 0.6, y = MCID_PP + 2.2, hjust = 0, size = GEOM_PT,
           label = sprintf("average: %.1f pp", MCID_PP)) +
  annotate("text", x = 0.6, y = thr$mcid_pp_naive + 2.2, hjust = 0,
           size = GEOM_PT, colour = "grey35",
           label = sprintf("one 'typical' lot: %.1f pp", thr$mcid_pp_naive)) +
  expand_limits(y = 45) +
  labs(title = "C  The same threshold, lot by lot",
       subtitle = "a lot at the top or bottom of the curve shows a smaller gap",
       x = NULL, y = "Viability gap equal to a 3-fold shift (pp)")
```

```{r m1-threshold, fig.width=8, fig.height=4}
(p1a_prime | p1b)
```

```{r mod1_dataset}
mod1_dataset <- sim$mcid_by_donor
colnames(mod1_dataset) <- c("Donor", "Readout_CPD-A", "Readout_CPD-B", "Difference")
gt::gt(mod1_dataset) %>%
  fmt_number(
    decimals = 1,
    use_seps = FALSE
  )
```

From Module III onwards the
comparison itself is made on the IC~50~ scale. Note in passing that translating
at a single "typical" lot would have given
`r round(thr$mcid_pp_naive, 1)` pp, over half again as large as the average. A
threshold has to be expressed on the same scale, and in the same units, as the
estimate it will be compared
with.

```{r m1-export}
export_data(sim$mcid_by_donor, "m01_relevance_threshold",
            "the 3-fold rule translated into a viability gap, lot by lot")
```

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Statistics cannot rescue an experiment whose question was never defined. The
threshold is a scientific decision, it is fixed before the data, and it must
be carried onto the scale you will actually measure.

---

# Module II — What types of variation are there?

### Story

The first lot comes off the plate reader. Six numbers. What can be said?

```{r m2-reveal}
one_lot <- viab %>%
  filter(donor_id == "HH1") %>%
  select(donor_id, compound, replicate, raw_rlu, viability_pct)

export_data(one_lot, "m02_one_lot",
            "six wells from a single hepatocyte lot, both compounds")
one_lot %>% dplyr::select(-raw_rlu)
```

### Show the points, not a bar

```{r m2-dots, fig.width=5, fig.height=3.4}
p2a <- one_lot %>%
  ggplot(aes(compound, viability_pct, colour = compound)) +
  stat_summary(fun = mean, geom = "col", fill = "grey85", colour = NA,
               width = 0.68) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.18,
               colour = "grey30") +
  tight_x() +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  labs(title = "Bar chart",
       x = NULL, y = "Viability (% of vehicle)") +
  coord_cartesian(ylim = c(0, 80))

p2b <- one_lot %>%
  ggplot(aes(compound, viability_pct, colour = compound)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.08, height = 0, seed = 1)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5,
               colour = "grey25", linewidth = 0.6) +
  tight_x() +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  labs(title = "Dot plot (strip plot)", subtitle = "Horizontal bars: mean values",
       x = NULL, y = NULL) +
  coord_cartesian(ylim = c(0, 80))

p2a | p2b
```

The bar chart is not wrong. It is simply compatible with almost any underlying
data, which is why it is a poor default.

```{r}
one_lot_stat <- one_lot %>%
  dplyr::select(compound, viability_pct) %>%
  group_by(compound) %>%
  summarise(mean_viability=mean(viability_pct),
            sd_viability=sd(viability_pct)) %>%
  mutate(CV=sd_viability/mean_viability)
one_lot_stat %>%
  gt %>%
    fmt_number(
    decimals = 2,
    use_seps = FALSE
  )
```

### The mean moves every time

Nothing about the lot or the compound changes below. The same measurement
process is repeated twenty times, three wells at a time.

```{r m2-mean-moves, fig.width=5, fig.height=3.5}
set.seed(101)  ## fixed so the deck and handout stay in sync
true_v <- 100 * hill_viability(CONC,
                               sim$truth$ic50_true[sim$truth$donor_id == "HH1" &
                                                   sim$truth$compound == "CPD-B"],
                               sim$params$hill_A)
tech_cv <- 0.05 + 0.12 * (1 - true_v / 100)

repeats <- tibble(experiment = rep(1:20, each = 3)) %>%
  mutate(well = rep(1:3, 20),
         viability = true_v * (1 + rnorm(n(), 0, tech_cv)))

rep_means <- repeats %>%
  group_by(experiment) %>%
  summarise(mean_viability = mean(viability), .groups = "drop")

ggplot(repeats, aes(experiment, viability)) +
  geom_hline(yintercept = true_v, linetype = "dashed", colour = "#c1443b",
             linewidth = 1) +
  geom_point(colour = "grey55", size = 1.9, alpha = 0.8) +
  geom_point(data = rep_means, aes(y = mean_viability), size = 2.9,
             colour = "#1f6fb4") +
  geom_line(data = rep_means, aes(y = mean_viability), colour = "#1f6fb4",
            linewidth = 0.9) +
  annotate("text", x = 20.4, y = true_v, hjust = 1, vjust = -0.8, size = GEOM_PT,
           colour = "#c1443b", label = "Ground truth") +
  labs(title = "20 experiments",
       subtitle = "Individual replicates (gray) and mean (blue), respectively",
       x = "Repeat of the experiment", y = "Viability (% of vehicle)")
```

```{r many_simulation, fig.height=4, fig.width=5}
many_repeats <- tibble(experiment = rep(1:1000, each = 3)) %>%
  mutate(well = rep(1:3, 1000),
         viability = true_v * (1 + rnorm(n(), 0, tech_cv)))
many_repeats_summary <- many_repeats %>%
  group_by(experiment) %>% summarise(mean_viability=mean(viability),
                                     sd_viability=sd(viability)) %>%
  mutate(CV=sd_viability/mean_viability)
{
  compactPar()
  hist(many_repeats_summary$mean_viability, breaks=100,
       xlab="Mean viability of three replicates", ylab="Frequency",
       col="lightblue",
       main = "Histogram of sample mean in 1000 simulations")
  abline(v=true_v, lty=2, lwd=2, col="#B2182B")
  legend("topleft", 
         legend=sprintf("HH1 true value: %.1f\nAvg. CV:%.1f%%", true_v, tech_cv*100), bty="n")
}
```

```{r many_simulation_CV, fig.height=4, fig.width=5}
{
  compactPar()
  hist(many_repeats_summary$CV, breaks=100,
       xlab="CV of three replicates", ylab="Frequency",
       col="mistyrose",
       main = "Histogram of CV in 1000 simulations")
  abline(v=tech_cv, lty=2, lwd=2, col="#2B3fB3")
  legend("topright", 
         legend=sprintf("True CV:%.1f%%", tech_cv*100), bty="n")
}
```

```{r many_more_simulation, fig.height=8, fig.width=8}
many_more_repeats <- tibble(experiment = rep(1:1000, each = 8)) %>%
  mutate(well = rep(1:8, 1000),
         viability = true_v * (1 + rnorm(n(), 0, tech_cv)))
many_more_repeats_summary <- many_more_repeats %>%
  group_by(experiment) %>% summarise(mean_viability=mean(viability),
                                     sd_viability=sd(viability)) %>%
  mutate(CV=sd_viability/mean_viability)
{
  compactPar(mfrow=c(2,2))
  
  hist(many_repeats_summary$mean_viability, breaks=100,
       xlab="Mean viability of three replicates", ylab="Frequency",
       col="lightblue", xlim=c(35,50),
       main = "Histogram of sample mean in 1000 simulations")
  abline(v=true_v, lty=2, lwd=2, col="#B2182B")
  legend("topleft", 
         legend=sprintf("HH1 true value: %.1f\nAvg. CV:%.1f%%", true_v, tech_cv*100), bty="n")
  
    hist(many_repeats_summary$CV, breaks=100,
       xlab="CV of three replicates", ylab="Frequency",
       col="mistyrose", xlim=c(0, 0.35),
       main = "Histogram of CV in 1000 simulations")
  abline(v=tech_cv, lty=2, lwd=2, col="#2B3fB3")
  legend("topright", 
         legend=sprintf("True CV:%.1f%%", tech_cv*100), bty="n")
  

  hist(many_more_repeats_summary$mean_viability, breaks=100,
       xlab="Mean viability of eight replicates", ylab="Frequency",
       col="skyblue",  xlim=c(35,50),
       main = "Histogram of sample mean in 1000 simulations")
  abline(v=true_v, lty=2, lwd=2, col="#B2182B")
  legend("topleft", 
         legend=sprintf("HH1 true value: %.1f\nAvg. CV:%.1f%%", true_v, tech_cv*100), bty="n")
  
    hist(many_more_repeats_summary$CV, breaks=100,
       xlab="CV of eight replicates", ylab="Frequency",
       col="#FFDECC", xlim=c(0, 0.35), lwd=0.05,
       main = "Histogram of CV in 1000 simulations")
  abline(v=tech_cv, lty=2, lwd=2, col="#2B3fB3")
  legend("topright", 
         legend=sprintf("True CV:%.1f%%", tech_cv*100), bty="n")
}
```


The truth is a fixed line. Every mean sits somewhere else. That gap is
**sampling uncertainty**, and it exists even when the experiment is perfect.
A confidence interval, later, is nothing more than an attempt to say how wide
this cloud is — from a single experiment.


### Not all variation is the same

```{r m2-taxonomy, fig.width=5, fig.height=3.7}
set.seed(102)  ## fixed so the deck and handout stay in sync
n_lots <- 8
base   <- 100 * hill_viability(CONC, sim$params$ic50_A, sim$params$hill_A)
lot_shift <- 100 * hill_viability(
  CONC, sim$params$ic50_A * 10^rnorm(n_lots, 0, log10(sim$params$donor_gsd)),
  sim$params$hill_A)

taxonomy <- bind_rows(
  tibble(layer = "1. Technical",
         value = base * (1 + rnorm(n_lots * 3, 0, 0.06))),
  tibble(layer = "2. + Donor effect",
         value = rep(lot_shift, each = 3) * (1 + rnorm(n_lots * 3, 0, 0.06))),
  tibble(layer = "3. + Batch effect",
         value = rep(lot_shift, each = 3) * (1 + rnorm(n_lots * 3, 0, 0.06)) *
                 rep(c(1, 1 / 1.2), each = n_lots * 3 / 2))
) %>%
  mutate(layer = ofactor(layer))

ggplot(taxonomy, aes(layer, value, colour = layer)) +
  geom_point(position = position_jitter(width = 0.14, seed = 3),
             size = 2.2, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.58, colour = "grey25") +
  tight_x() +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  labs(title = "Sources of variation and their impact on readout",
       x = NULL, y = "Viability (% of vehicle)")
```

Four kinds, and they call for four different responses:

| Kind | Example here | What to do |
|---|---|---|
| **Wanted** | the compound effect | measure it |
| **Unavoidable** | well-to-well pipetting | replicate, then average |
| **Controllable** | run day, plate, operator | block it, or standardize it |
| **Confounding** | run day that tracks compound | redesign — it cannot be fixed later |

### Bias and variance

```{r m2-target, fig.width=4, fig.height=4}
set.seed(103)  ## fixed so the deck and handout stay in sync
target_data <- expand_grid(
    bias     = c("unbiased", "biased"),
    variance = c("low variance", "high variance")) %>%
  mutate(shots = pmap(list(bias, variance), function(b, v) {
    ctr <- if (b == "biased") c(1.45, 1.05) else c(0, 0)
    s   <- if (v == "high variance") 1.15 else 0.32
    tibble(x = rnorm(24, ctr[1], s), y = rnorm(24, ctr[2], s))
  })) %>%
  unnest(shots) %>%
  mutate(bias     = factor(bias, levels = c("unbiased", "biased")),
         variance = factor(variance, levels = c("low variance", "high variance")))

## target rings, drawn as polygons so no extra package is needed
rings <- expand_grid(r = c(0.6, 1.2, 1.8, 2.4),
                     theta = seq(0, 2 * pi, length.out = 120)) %>%
  mutate(x = r * cos(theta), y = r * sin(theta))

ggplot(target_data) +
  geom_path(data = rings, aes(x, y, group = r), colour = "grey75") +
  geom_point(aes(x, y), colour = "#c1443b", size = 1.7, alpha = 0.85) +
  geom_point(aes(x = 0, y = 0), colour = "grey40", size = 1.6, shape = 3) +
  facet_grid(variance ~ bias) +
  coord_equal(xlim = c(-2.55, 2.55), ylim = c(-2.55, 2.55), expand = FALSE) +
  labs(title = "Bias and variance",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

Two scenarios to argue about:

* **Scenario A.** Every control sample is pipetted with pipette A, every
  treated sample with pipette B.
* **Scenario B.** Within one experiment, every replicate uses a different
  consumable batch, "to make it representative".

> Which one makes it hard to know whether the treatment caused the difference?
> Which one makes a real effect harder to detect?

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Variation is not a sign of a bad experiment. Some of it is the signal, some is
the price of measuring, some is a design choice you still have, and some is a
design mistake you no longer have.

---

# Module III — What are biological and technical replicates, and what are their uses?

### Dataset reveal

All 48 measurements are now on the table.

```{r m3-reveal}
export_data(viab %>% select(donor_id, donor_sex, donor_age, compound, weekday,
                            plate_id, well, replicate, conc_uM, raw_rlu,
                            viability_pct),
            "m03_all_wells_long",
            "the full 48-well dataset, one row per well (long format)")

viab %>%
  select(donor_id, compound, weekday, replicate, viability_pct) %>%
  head(12)
```

### The whole dataset

Every number the analysis uses, in three steps: the 48 wells as measured, the
16 lot-level summaries they average to, and the 8 within-lot fold shifts that
are the quantity we actually decide on.

```{r m3-table-wells}
viab_viz <- viab %>%
  transmute(Lot = donor_id, Compound = compound, Day = weekday,
            Well = replicate,
            ##`Raw RLU` = raw_rlu,
            `Viability (%)` = viability_pct,
            `IC50 (µM)` = ic50_uM)
viab_viz %>%
  knitr::kable(align = "llcrrrr",
               caption = "Data of 48 wells. Viability is the output of the plate reader at one particular concentration of both compounds. IC50 is derived from fitting a Hill function to the concentration-response data.") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                            full_width = FALSE, font_size = 12) %>%
  kableExtra::scroll_box(height = "360px")
```

```{r}
viab_viz %>% 
  dplyr::select(-Day, -Well, -`IC50 (µM)`) %>%
  pivot_wider(names_from=c("Lot"),
              values_from="Viability (%)",
              values_fn=~paste(.x, collapse=", ")) %>%
  t
```

```{r m3-table-lots}
lot_tab <- dm %>%
  transmute(Lot = donor_id, #Sex = donor_sex, Age = donor_age, Day = weekday,
            Compound = compound, ##Wells = n_wells,
            `Mean viability (%)` = round(viability, 1),
            `SD` = round(sd_wells, 1),
            `CV (%)`=round(`SD`/`Mean viability (%)`*100, 1),
            `IC50 (µM)` = round(ic50_uM, 2))

knitr::kable(lot_tab, align = "llrrrr",
             caption = "The 16 lot-compound groups") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                            full_width = FALSE, font_size = 12)
```

```{r m3-table-folds}
folds %>%
  transmute(Lot = donor_id,
            `CPD-A IC50 (µM)` = round(`CPD-A`, 2),
            `CPD-B IC50 (µM)` = round(`CPD-B`, 2),
            `Fold shift` = round(fold, 2)) %>%
  knitr::kable(align = "lrrr",
               caption = "The 8 lot groups") %>%
  kableExtra::kable_styling(bootstrap_options = c("striped", "condensed"),
                            full_width = FALSE, font_size = 12)
```

> We have 48 numbers. What is *n*? 48? 16? 8? Something else?

### Two kinds of replicate

**Technical replicate** — a repeated measurement made to characterize the
*measurement* process. Three wells from one lot tell you how reproducibly you
can pipette and read that lot.

**Biological (independent) replicate** — an independently sampled unit that
carries new information about the *population*. A second hepatocyte lot is a
second donor, a second liver, a second draw from the population you want to
speak about.

Three wells from one lot are not three lots. Averaging the three wells gives
you a better measurement of **that one lot**; it does not give you a second
lot.

Here, the independent experimental unit is the **hepatocyte lot**: *n* = 8.

### Pseudoreplication, made visible

The same comparison, run twice. Nothing changes except what is treated as
independent.

```{r m3-pseudoreplication}
## From here on the comparison is made on the IC50 scale, in the log, where
## the effect is additive and the lot effect cancels. Both analyses below ask
## the same question; they differ only in what they count as independent.
wrong <- t.test(log10(ic50_uM) ~ compound, data = viab)   # 48 wells
right <- t.test(log10(ic50_uM) ~ compound, data = dm)     # 8 lots

## both are CPD-A minus CPD-B; flip to CPD-B / CPD-A and back to folds
flip <- function(tt, lab, n_per_group)
  tibble(analysis = lab, n_per_group = n_per_group,
         fold = 10^-unname(diff(rev(tt$estimate))),
         ci_low = 10^-tt$conf.int[2], ci_high = 10^-tt$conf.int[1],
         p_value = tt$p.value)

pseudo <- bind_rows(
  flip(wrong, "wells treated as independent (n = 48)", 24),
  flip(right, "lots treated as independent (n = 8)",    8)) %>%
  mutate(ci_ratio = ci_high / ci_low,
         across(c(fold, ci_low, ci_high, ci_ratio), ~ round(.x, 2)),
         p_value = signif(p_value, 3))

export_data(pseudo, "m03_pseudoreplication",
            "the same comparison analyzed at the well level and at the lot level")
pseudo
```

```{r m3-pseudo-fig, fig.width=7, fig.height=3.2}
pseudo %>%
  transmute(label = c("48 wells as 24 comparisons", "8 lots as 8 comparisons"),
            estimate = fold, lwr = ci_low, upr = ci_high) %>%
  forest("Pseudoreplicate versus independent replicate",
         "Treating pseudoreplicate as biological replicate inflates the results")
```

The point estimate is identical to the second decimal:
`r sprintf("%.2f", pseudo$fold[1])`-fold either way. The p-value is about
`r round(pseudo$p_value[2] / pseudo$p_value[1])` times **smaller** when wells
are counted as independent, and the interval is materially narrower. Counting
wells adds no information about lots; it only makes the software believe there
is more of it. The extra confidence is manufactured.

And this time the shortcut does not get away with it. The wells-as-independent
interval runs from `r pseudo$ci_low[1]` to `r pseudo$ci_high[1]`-fold, which
**straddles the 3-fold threshold** and reads *inconclusive*. The correct
analysis in Module V will give `r sprintf("%.2f", fold_test(folds)$lwr)` to
`r sprintf("%.2f", fold_test(folds)$upr)`-fold — a clear *act*.

Two errors sit on top of each other here, and it is worth separating them.
Counting wells inflates *n* and **narrows** the interval; ignoring the pairing
leaves the lot-to-lot variation in the error term and **widens** it. The
well-level analysis does both, so its interval is not reliably too narrow or
too wide — it is simply not entitled to whatever width it happens to have.

### Twelve measurements, two ways to spend them

```{r m3-designs, fig.width=5.5, fig.height=5.5}
set.seed(104)  ## fixed so the deck and handout stay in sync
gp <- sim$params

## Experiment A: 12 lots x 1 well   |   Experiment B: 3 lots x 4 wells
one_study <- function(n_lots, n_wells) {
  ic50 <- gp$ic50_A * 10^rnorm(n_lots, 0, log10(gp$donor_gsd))
  v    <- 100 * hill_viability(gp$screen_conc, ic50, gp$hill_A)
  cv   <- 0.05 + 0.12 * (1 - v / 100)
  obs  <- v * (1 + matrix(rnorm(n_lots * n_wells), nrow = n_lots) * cv)
  ## on the IC50 scale, in the log, to match the rest of the module
  lg   <- log10(ic50_from_viability(obs, gp$screen_conc, gp$hill_A))
  tibble(
    ## how well can we see how much lots differ from each other?
    sd_between_lots = if (n_lots > 1) sd(rowMeans(lg)) else NA_real_,
    ## how well can we see how reproducible the measurement is?
    sd_within_lot   = if (n_wells > 1) mean(apply(lg, 1, sd)) else NA_real_)
}

design_cmp <- bind_rows(
  map_dfr(1:params$n_sim, ~ one_study(12, 1)) %>%
    mutate(design = "A: 12 donors × 1 rep"),
  map_dfr(1:params$n_sim, ~ one_study(6, 2)) %>%
    mutate(design = "B: 6 donors × 2 reps"),
  map_dfr(1:params$n_sim, ~ one_study(4, 3)) %>%
    mutate(design = "C: 4 donors × 3 reps"),
  map_dfr(1:params$n_sim, ~ one_study(3, 4)) %>%
    mutate(design = "D: 3 donors × 4 reps"),
  map_dfr(1:params$n_sim, ~ one_study(2, 6)) %>%
    mutate(design = "E: 2 donors × 6 reps"),
  map_dfr(1:params$n_sim, ~ one_study(1, 12)) %>%
    mutate(design = "F: 1 donors × 12 reps"))
  

export_data(design_cmp %>% group_by(design) %>%
              summarise(across(everything(),
                               list(median = ~ median(.x, na.rm = TRUE),
                                    iqr = ~ IQR(.x, na.rm = TRUE)))),
            "m03_design_comparison",
            "12 measurements spent two ways: what each design can estimate")

design_cmp %>%
  pivot_longer(starts_with("sd_"), names_to = "quantity", values_to = "estimate") %>%
  mutate(quantity = recode(quantity,
           sd_between_lots = "SD between donors \n(biological variability)",
           sd_within_lot   = "SD within donor \n(technical variability)")) %>%
  ggplot(aes(design, estimate, fill = design)) +
  geom_violin(alpha = 0.6, colour = NA, na.rm = TRUE, width = 0.92) +
  geom_boxplot(width = 0.13, outlier.shape = NA, na.rm = TRUE) +
  tight_x() +
  facet_wrap(~ quantity, scales = "free_y") +
  scale_fill_manual(values = royalbluegrayred(6), guide = "none") +
  theme(axis.text.x = element_text(angle=90, hjust=1)) +
  labs(title = "How to balance biological and technical replicates?",
       subtitle = "Results from thousands of simulations. Empty panel: non-estimtable",
       x = NULL, y = expression("Estimated standard deviation (SD) of log"[10]~"IC"[50]))
```

Twelve lots × one well estimates lot-to-lot variation well and technical
variation not at all. Three lots × four wells does the opposite, and estimates
lot-to-lot spread from three points — which is barely an estimate. Neither is
better; they answer different questions. If the question is about *hepatocyte
lots in general*, spend the measurements on lots.

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Variation among biological replicates tells you about the population.
Variation among technical replicates tells you about your hands and your
instrument. *n* is the number of independent units, and the analysis has to
be done at that level.

---

# Module IV — What are randomization and blocking, and why are they important for experimental design?

### Story

Sixteen plates cannot be run in one sitting, so the assay is split over Friday
and Monday. Suppose the scheduler had done the natural thing and run **all
CPD-A plates on Friday and all CPD-B plates on Monday**.

Something real happened on Friday: the vehicle wells read about
`r scales::percent(sim$params$friday_vehicle_gain)` higher. Because viability is
expressed relative to each plate's own vehicle, a shift in the reference
propagates into every normalized value on that plate.

```{r m4-vehicle, fig.width=4.4, fig.height=3.4}
vehicle_wells <- sim$ctg %>%
  filter(design == "Blocked", well_type == "vehicle") %>%
  select(donor_id, compound, weekday, replicate, raw_rlu)

export_data(vehicle_wells, "m04_vehicle_wells",
            "vehicle (untreated) wells only - where the run-day effect lives")

vehicle_wells %>%
  ggplot(aes(weekday, raw_rlu / 1e6, fill = weekday)) +
  geom_boxplot(width = 0.52, alpha = 0.65, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1, seed = 2), size = 1.8,
             alpha = 0.7) +
  scale_x_discrete(expand = expansion(add = 0.38)) +
  scale_fill_manual(values = day_cols, guide = "none") +
  labs(title = "Background signal in vehicle wells",
       x = NULL, y = expression("Raw signal (10"^6~"RLU)"))
```

### The same lots, two schedules

```{r m4-designs, fig.width=6, fig.height=3.8}
sim$schedule %>%
  select(donor_id, compound, Blocked = weekday_blocked,
         Confounded = weekday_confounded) %>%
  pivot_longer(c(Blocked, Confounded), names_to = "design",
               values_to = "weekday") %>%
  export_data("m04_schedules", "run-day allocation under the two designs") %>%
  ggplot(aes(compound, fct_rev(donor_id), fill = weekday)) +
  geom_tile(colour = "white", linewidth = 1.2) +
  geom_text(aes(label = str_sub(weekday, 1, 3)), colour = "white",
            fontface = "bold", size = GEOM_PT) +
  facet_wrap(~ design) +
  scale_x_discrete(expand = flush) +
  scale_y_discrete(expand = flush) +
  scale_fill_manual(values = day_cols) +
  labs(title = "Day as a blocking factor or as a confounder",
       x = NULL, y = "Hepatocyte lot") +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

### What each design concludes

```{r m4-consequences}
design_result <- bind_rows(lot_fold(viab)          %>% mutate(design = "Blocked"),
                           lot_fold(sim$viab_conf) %>% mutate(design = "Confounded"))

## what we would see with no day effect and no measurement noise
true_fold <- sim$truth %>%
  select(donor_id, compound, ic50_true) %>%
  pivot_wider(names_from = compound, values_from = ic50_true) %>%
  summarise(x = exp(mean(log(`CPD-B` / `CPD-A`)))) %>% pull(x)

design_summary <- design_result %>%
  group_by(design) %>%
  summarise(fold_shift = exp(mean(log(fold))), .groups = "drop") %>%
  mutate(simulated_truth = true_fold,
         error_fold      = fold_shift / true_fold,
         across(where(is.numeric), ~ round(.x, 2)))

export_data(design_result, "m04_blocked_vs_confounded",
            "per-lot compound difference under both schedules")
design_summary
```

Both schedules land within a fold or so of the truth, and on this dataset the
confounded one is not obviously worse. That is the difficulty with
confounding: it does not announce itself.

### The experiment that makes it obvious

So run the experiment that settles it. Simulate the same study with the two
compounds made **exactly identical** — same IC~50~, same slope — and leave
everything else alone, including the Friday vehicle shift. Any difference the
analysis now reports is, by construction, not a compound difference.

```{r m4-null-compounds}
null_cpd <- simulate_workshop_data(seed = params$seed,
                                   ic50_B = sim$params$ic50_A)

test_pair <- function(d, label) {
  fold_test(lot_fold(d)) %>%
    transmute(schedule = label, fold_shift = estimate,
              ci_low = lwr, ci_high = upr, p_value)
}

null_result <- bind_rows(
  test_pair(null_cpd$viab,      "Blocked"),
  test_pair(null_cpd$viab_conf, "Confounded")) %>%
  mutate(verdict = if_else(p_value < 0.05,
                           "a 'significant' difference between identical compounds",
                           "correctly finds nothing"),
         across(c(fold_shift, ci_low, ci_high), ~ round(.x, 2)),
         p_value = signif(p_value, 3))

export_data(null_result, "m04_identical_compounds",
            "two IDENTICAL compounds run under both schedules")
null_result
```

```{r confounding_effects, fig.height=3, fig.width=4}
block_data <- null_cpd$viab %>%
  dplyr::select(donor_id, compound, ic50_uM) %>%
  group_by(donor_id, compound) %>%
  summarise(ic50_uM=mean(ic50_uM))
conf_data <- null_cpd$viab_conf %>%
  dplyr::select(donor_id, compound, ic50_uM) %>%
  group_by(donor_id, compound) %>%
  summarise(ic50_uM=mean(ic50_uM))
block_conf_data <- rbind(cbind(block_data, type="Blocking"),
                         cbind(conf_data, type="Confounding")) %>%
  mutate(compound=ifelse(compound=="CPD-B", "CPD-A'", "CPD-A"))
ggplot(block_conf_data, aes(x=compound, y=ic50_uM, color=compound)) +
  facet_wrap(~type) +
  geom_boxplot() +
  scale_colour_manual(values = c("CPD-A"="#1f6fb4",
                                 "CPD-A'"="#1f0fb4"), guide = "none") +
  geom_point() + geom_path(aes(group=donor_id), col="darkgray")
```

```{r m4-null-fig, fig.width=5, fig.height=3.0}
## the question here is "any difference at all?", so this one is read against
## zero rather than against the relevance threshold
null_result %>%
  mutate(row = factor(schedule, levels = c("Confounded", "Blocked")),
         detected = if_else(p_value < 0.05, "signficant difference",
                                            "non-significant")) %>%
  ggplot(aes(fold_shift, row, colour = detected)) +
  geom_vline(xintercept = 1, linewidth = 0.9, colour = "grey35") +
  scale_x_log10() +
  geom_linerange(aes(xmin = ci_low, xmax = ci_high), linewidth = 1.3) +
  geom_point(size = 2.8) +
  scale_colour_manual(values = c("signficant difference" = "#a33",
                                 "non-significant" = "#3b7a57")) +
  scale_y_discrete(expand = expansion(add = 0.5)) +
  labs(title = "Simulating a comparison of two identical compounds",
       subtitle = "An interval that does not include 1 is an artifact",
       x = "Apparent fold change in cytotoxicity IC50", y = NULL) +
  theme(legend.position = "bottom")
```

The blocked schedule reports nothing, correctly. The confounded schedule
reports a shift of
`r null_result$fold_shift[2]`-fold with *p* =
`r null_result$p_value[2]` — statistically significant, perfectly
reproducible, and entirely an artifact of which day the plates ran.

Nothing in the confounded dataset can reveal this. Day and compound are the
same column: every CPD-A plate is a Friday plate. No test, no model and no
amount of care in the analysis can separate two factors that never varied
independently. In the blocked schedule day varies *within* each compound, so
it becomes a nuisance you can estimate rather than one you cannot see.

> One subtlety, now much less awkward than it used to be. The day effect is
> **multiplicative** — the vehicle reference moves by a factor — and so is the
> quantity we now report. Working in the log of the IC~50~ is what lets a
> blocked day effect cancel in the contrast instead of merely shrinking it.
> Blocking
> makes the day effect cancel exactly in a *ratio*, but only approximately in
> a difference. Choosing a scale on which your effect is additive is the same
> decision we made in Module I, and it keeps repeating.

### The three levers

| Lever | Use it for | Effect | Failure mode |
|---|---|---|---|
| **Block** | known, structural sources: day, plate, operator | removes confounding | blocking on something that tracks treatment does nothing |
| **Randomize** | everything you cannot block or foresee: well position, order | converts unknown bias into honest variance | small experiments can still randomize into an imbalance |
| **Standardize** | irrelevant sources: same pipette, one consumable lot | shrinks variance | standardizing *per arm* manufactures bias |

The last column is where the damage is done. Using the same pipette throughout
reduces variance. Using pipette A for condition A and pipette B for condition B
is *also* "standardizing", and it creates a confounder. Deliberately varying
consumable batch within every small experiment "to be representative" is not
robustness — it is added noise that makes a real effect harder to see.

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Good design decides which variation the statistics will be able to separate
later. After the experiment, no analysis can unmix a confounder.

---

# Module V — How do we compare two groups?

### Story

Here is the study as two clouds of dots — the way this comparison is usually
drawn.

```{r m5-clouds, fig.width=5, fig.height=3.7}
set.seed(105)  ## fixed so the deck and handout stay in sync
p5a <- dm %>%
  ggplot(aes(compound, viability, colour = compound)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.09, height = 0, seed = 7)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "grey25") +
  tight_x() +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "Two groups, unpaired",
       x = NULL, y = "Viability (% of vehicle)")

p5b <- dm %>%
  ggplot(aes(compound, viability)) +
  geom_line(aes(group = donor_id), colour = "grey65", linewidth = 0.8) +
  geom_point(aes(colour = compound), size = 2.8, alpha = 0.9) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "grey25") +

  scale_x_discrete(expand = expansion(add = 0.3)) +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "Two groups, paired",
       x = NULL, y = NULL)

p5a | p5b
```

> What information did the first graph hide?

Every lot was tested with **both** compounds. The vertical scatter in panel A
is mostly lots being different from one another, not compounds being similar.
Once the lines are drawn, the lot differences stop competing with the signal.

### Four tests, one dataset

```{r m5-tests}
## everything on log10 IC50, where the effect is additive, reported as folds
wide <- folds %>% select(donor_id, `CPD-A`, `CPD-B`)
lg   <- dm %>% mutate(log_ic50 = log10(ic50_uM))

tests <- list(
  "Student t, unpaired (equal variance assumed)" =
    t.test(log_ic50 ~ compound, data = lg, var.equal = TRUE),
  "Welch t, unpaired (no equal-variance assumption)" =
    t.test(log_ic50 ~ compound, data = lg),
  "Paired t" =
    t.test(log10(wide$`CPD-B`), log10(wide$`CPD-A`), paired = TRUE),
  "Wilcoxon signed rank (paired, non-parametric)" =
    wilcox.test(log10(wide$`CPD-B`), log10(wide$`CPD-A`),
                paired = TRUE, conf.int = TRUE))

test_table <- imap_dfr(tests, function(tt, nm) {
  est <- if (grepl("^Paired|^Wilcoxon", nm)) unname(tt$estimate)
         else unname(-diff(rev(tt$estimate)))
  lo  <- tt$conf.int[1] * if (grepl("unpaired", nm)) -1 else 1
  hi  <- tt$conf.int[2] * if (grepl("unpaired", nm)) -1 else 1
  tibble(test = nm, fold_shift = 10^est,
         ci_low = 10^min(lo, hi), ci_high = 10^max(lo, hi),
         p_value = tt$p.value)
}) %>%
  mutate(ci_ratio = ci_high / ci_low,
         beats_threshold = if_else(ci_low > MCID_FOLD, "yes", "inconclusive"),
         across(c(fold_shift, ci_low, ci_high, ci_ratio), ~ round(.x, 2)),
         p_value = signif(p_value, 3))

export_data(test_table, "m05_four_tests",
            "the same comparison under four tests")
test_table
```

```{r m5-tests-fig, fig.width=6.8, fig.height=3}
test_table %>%
  transmute(label = c("Student t, unpaired", "Welch t, unpaired",
                      "Paired t", "Wilcoxon signed rank"),
            estimate = fold_shift, lwr = ci_low, upr = ci_high) %>%
  forest("Comparing between unpaired and paired tests",
         sprintf("grey: shifts too small to act on (threshold %g-fold)",
                 MCID_FOLD))
```

Read the last two columns, not the p-value column. Every test agrees there is
a difference. They do **not** agree on whether it clears the
`r MCID_FOLD`-fold threshold the team committed to in Module I: the
unpaired analyses cannot tell, the paired analysis can. Ignoring the pairing
did not just cost significance — it cost the decision.

### The decision rule

Are observations linked by the experimental design?

* same lot under both compounds → **paired**
* same animal before and after → **paired**
* separately sampled groups → **unpaired**

Pairing is a property of the design, fixed before the data exist. It is never
something you discover by trying both and keeping the better result.

If the groups really are independent, do not assume equal variances without a
reason. **Welch's t-test** does not require equal variances, and when the
variances happen to be equal it costs almost nothing — which is exactly what
the first two rows of the table show here. On the scale the test is run on, the
lot-level SDs of log~10~ IC~50~ are
`r paste(round(tapply(log10(dm$ic50_uM), dm$compound, sd), 2), collapse = " and ")`,
close enough that Student and Welch agree to the second decimal. That is the
argument *for* Welch as the default: you get the protection when you need it
and pay nothing when you do not, and you avoid having to test for equal
variances first — a pre-test that itself has poor power at these sample sizes.

The Wilcoxon signed-rank test is the paired non-parametric option. On eight
pairs it gives a similar estimate with a larger p-value, because ranking eight
numbers throws away most of the information they contain. With *n* = 8 the
smallest p-value it can ever return is 1/128 ≈ 0.008 — so it cannot produce
strong evidence here no matter what the data look like. Non-parametric is not
a free lunch; it buys robustness to distributional assumptions with power.

### Beyond two groups

Two compounds in the same lots is the simplest comparison there is, and it is
the one this workshop stays with. Real projects outgrow it quickly, and the
tools that follow are extensions of the same idea rather than new subjects:
**ANOVA** when there are more than two groups, and **linear regression** when
you need to adjust for something you measured but did not control — run day,
passage number, donor age. Both still require you to know what the independent
unit is and which observations are linked, so every question in this module
still has to be answered first.

### Visualization hierarchy

1. the raw observations
2. the structure — pairing, blocks, batches
3. the estimated effect
4. its uncertainty

Bars, SEM whiskers and asterisks come after all four, if at all.

### Wide and long, the same 16 numbers

```{r m5-wide-long}
export_data(dm %>% select(donor_id, donor_sex, donor_age, weekday, compound,
                          viability, sd_wells),
            "m05_lot_means_long",
            "lot-level means, long format - one row per lot x compound (for R/Python)")
export_data(wide, "m05_lot_means_wide",
            "lot-level means, wide format - one row per lot (for Excel/Prism)")

wide %>% mutate(across(where(is.numeric), ~ round(.x, 1)))
```

The **long** format has one row per measurement and a column naming each
factor; it is what `ggplot2`, `dplyr` and most statistical models expect. The
**wide** format spreads one factor across columns; it is what spreadsheets and
Prism are built for, and it is the right shape for a paired test because the
pairing is carried by the row. Both files are exported. Converting between
them is `pivot_wider()` / `pivot_longer()`, and the conversion is lossless
only if the long table carries every identifier — which is a good reason to
keep lot, day and plate as real columns rather than as sheet names.

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

The statistical test follows the experimental design. Draw the structure
before you test it, and the right test is usually obvious.

---

# Module VI — How do we detect and handle outliers?

### Story

Two values in the dataset look wrong. They are not the same kind of thing.

```{r m6-candidates, fig.width=6, fig.height=3.5}
lot_flags <- viab %>%
  group_by(donor_id, compound) %>%
  mutate(well_z = (viability_pct - median(viability_pct)) /
                  (mad(viability_pct) + 1e-9)) %>%
  ungroup()

export_data(lot_flags %>% select(donor_id, compound, weekday, replicate,
                                 viability_pct, well_z),
            "m06_outlier_candidates",
            "all wells with a within-condition robust z score")

lot_flags %>%
  ggplot(aes(fct_reorder(donor_id, viability_pct), viability_pct,
             colour = compound)) +
  geom_point(size = 2.4, alpha = 0.85,
             position = position_dodge(width = 0.55)) +
  scale_x_discrete(expand = expansion(add = 0.5)) +
  ## the bubble well is the CPD-B point of HH2, dodged to the right of center
  ## annotate("curve", x = 5.3, xend = 4.3, y = 33, yend = 20.3, curvature = -0.3,
  ##         arrow = arrow(length = unit(0.18, "cm")), colour = "grey35") +
  ## annotate("text", x = 7, y = 40, hjust = 1, size = GEOM_PT, colour = "grey25",
  ##         label = "Suspicious: one well,\nfar from other two replicates") +
  ## annotate("curve", x = 1.6, xend = 0.92, y = 66, yend = 45, curvature = 0.3,
  ##          arrow = arrow(length = unit(0.18, "cm")), colour = "grey35") +
  ## annotate("text", x = 1.7, y = 72, hjust = 0, size = GEOM_PT, colour = "grey25",
  ##         label = "candidate 2: a whole lot,\nlow for BOTH compounds") +
  scale_colour_manual(values = cpd_cols) +
  labs(title = "Data of 48 wells",
       x = "Hepatocyte lot", y = "Viability (% of vehicle)") +
  coord_cartesian(ylim = c(0, 104)) +
  theme(legend.position = "bottom")
```

### Candidate 1 — a single well

```{r m6-well}
lot_flags %>%
  filter(donor_id == sim$params$bubble_donor,
         compound == sim$params$bubble_compound) %>%
  select(donor_id, compound, replicate, raw_rlu, viability_pct, well_z)
```

The laboratory notebook for that plate says:

> *Air bubble observed in well B; pipette felt strange.*

An independent, documented, technical cause, recorded **before** anyone looked
at the result. The other two wells of the same condition agree with each
other. This is a measurement failure, and excluding it is a QC decision, not
a statistical one.

### Candidate 2 — a whole lot

```{r m6-grubbs}
grubbs <- function(x) {
  x <- x[!is.na(x)]; n <- length(x)
  G     <- max(abs(x - mean(x))) / sd(x)
  tcrit <- qt(0.05 / (2 * n), df = n - 2)
  Gcrit <- (n - 1) / sqrt(n) * sqrt(tcrit^2 / (n - 2 + tcrit^2))
  tibble(n = n, G = round(G, 3), G_crit = round(Gcrit, 3), flagged = G > Gcrit)
}

dm %>%
  group_by(compound) %>%
  group_modify(~ grubbs(.x$viability)) %>%
  ungroup()
```

Lot `r sim$params$sensitive_donor` is more sensitive to **both** compounds —
its IC~50~ is genuinely about
`r sim$params$sensitive_fold`-fold lower. A routine outlier test will flag it
anyway, because it is answering the wrong question: Grubbs asks whether a value
is improbable under an assumed distribution, never whether it is *wrong*.
Nothing in the QC record marks that plate. A lot that is genuinely more
sensitive is precisely the observation a safety margin should be built on.

### A workflow, not a test

1. **Verify** — transcription, instrument, pipetting, sample identity
2. **Investigate** — is there an experimental explanation in the record?
3. **Apply pre-defined rules** — was an exclusion criterion fixed before the
   result was seen?
4. **Do not delete because it is inconvenient** — biology is often surprising
5. **Run a sensitivity analysis** — does the conclusion change either way?
6. **Document** — report what was excluded and why

### Step 5, done

```{r m6-sensitivity}
all_wells <- viab
no_bubble <- viab %>% filter(!bubble)
drop_lot  <- function(d) d %>% filter(donor_id != sim$params$sensitive_donor)

run_paired <- function(d, label) {
  w <- lot_fold(d)
  fold_test(w) %>%
    transmute(scenario = label, n_lots = nrow(w), fold_shift = estimate,
              ci_low = lwr, ci_high = upr, p_value)
}

sensitivity <- bind_rows(
  run_paired(all_wells,            "keep everything"),
  run_paired(no_bubble,            "exclude the air-bubble well (documented QC failure)"),
  run_paired(drop_lot(all_wells),  "exclude the sensitive lot (no technical justification)"),
  run_paired(drop_lot(no_bubble),  "exclude both")) %>%
  mutate(clears_threshold = if_else(ci_low > MCID_FOLD, "yes", "inconclusive"),
         across(c(fold_shift, ci_low, ci_high), ~ round(.x, 2)),
         p_value = signif(p_value, 3))

export_data(sensitivity, "m06_sensitivity_analysis",
            "the conclusion under four exclusion scenarios")
sensitivity
```

```{r m6-sensitivity-fig, fig.width=6.8, fig.height=3.4}
sensitivity %>%
  transmute(label = c("keep everything",
                      "drop the air-bubble well",
                      "drop the most sensitive donor",
                      "drop both"),
            estimate = fold_shift, lwr = ci_low, upr = ci_high) %>%
  forest("Sensitivity analysis")
```

Each row is the estimated fold shift, with the range of shifts the data are
compatible with. We have not yet said what that range means — that is the next
module — but it already answers the question that matters here: **does the
answer depend on which values we remove?**

Dropping the air-bubble well does not change it. Dropping the sensitive lot
**does**: the interval widens past the 3-fold line and the verdict falls back
to *inconclusive*. The exclusion we could justify from the QC record costs
nothing; the exclusion we could not justify is the one that changes the
answer.

Dropping the bubble well changes the estimate barely at all — because it was
one of three wells that were then averaged, and averaging is already doing
most of the work. Dropping the sensitive lot is a different matter: the
pairing had absorbed that lot completely (it is low for both compounds, so its
*fold shift* is ordinary), so removing it discards a lot that was costing
nothing and carrying real information about how variable human hepatocytes
are. Note which way it moves the result: dropping it does not make the effect
look bigger, it makes the study less certain.

> The companion document extends this to a whole plate that fails
> pre-specified QC, with Z' factor, vehicle CV and signal window as the
> criteria, and to the case where you should model the extra variance rather
> than delete it.

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Ask which question you are answering: is the value *implausible given the
controls*, or merely *inconvenient given the hypothesis*? The first is a QC
exclusion. The second is data manipulation. Running it both ways is cheap and
almost always the right thing to report.

---

# Module VII — What do we mean by p-values, effect sizes, and confidence intervals?

### Story

Assume, for a moment, that the two compounds are **identical**. Then run the
whole study — 8 lots, 3 wells, both compounds — `r params$n_sim` times in that
imaginary world, and record the difference each time.

```{r m7-null, fig.width=6.8, fig.height=3.5}
set.seed(106)  ## fixed so the deck and handout stay in sync
null_world <- replicate_studies(params$n_sim, n_donors = 8, n_wells = 3,
                                fold = 1)
observed   <- fold_test(folds)$estimate        # a fold, e.g. 4.8x

export_data(null_world %>% select(sim, estimate, p_value),
            "m07_null_simulation",
            sprintf("%d simulated studies in a world where the compounds are identical",
                    params$n_sim))

## a fold shift is a ratio, so "at least as extreme as ours" means at least as
## far from 1 in the log - and the two dashed lines are 4.8x and 1/4.8x
p6a <- ggplot(null_world, aes(estimate)) +
  geom_histogram(bins = 45, fill = "grey78", colour = "white") +
  geom_histogram(data = filter(null_world,
                               abs(log10(estimate)) >= abs(log10(observed))),
                 bins = 45, fill = "#c1443b", colour = "white") +
  scale_x_log10(limits=c(0.3, 3)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04))) +
  labs(title = "Compare CPD-A and CPD-A'",
       subtitle = "With identical population mean",
       x = "Fold change between compounds", y = "Simulated studies")

p6b <- ggplot(null_world, aes(p_value)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "grey78", colour = "white") +
  geom_histogram(data = filter(null_world, p_value < 0.05),
                 breaks = seq(0, 1, 0.05), fill = "#c1443b", colour = "white") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.04))) +
  geom_hline(yintercept=params$n_sim*0.05, linetype=2) +
  labs(title = "P-values under the null hypothesis",
       subtitle = sprintf("%.1f%% simulations report p<0.05",
                          100 * mean(null_world$p_value < 0.05)),
       x = "p-value", y = "Simulated studies")

p6a | p6b
```

Panel A *is* the p-value: the red area is the fraction of studies in a
no-difference world that would produce a difference at least as extreme as
ours. Here that fraction is
`r sprintf("%.4f", mean(abs(log10(null_world$estimate)) >= abs(log10(observed))))`,
and the paired t-test reports
`r format.pval(tests[['Paired t']]$p.value, digits = 3)` — the same quantity, computed from
a formula instead of from `r params$n_sim` simulated experiments.

Panel B is worth a minute on its own. When there is genuinely nothing to find,
p-values are **uniform**: every value is as likely as every other, and 5% land
under 0.05. That is what a 5% false-positive rate means, and it is why a
single p-value below 0.05 is weak evidence on its own.

### What a p-value is not

A p-value is **not**:

* the probability that the compounds are identical
* the probability that the result happened by chance
* the probability that the experiment will replicate
* anything at all about the **size** of the effect

```{r effect_size_p_value, fig.height=4, fig.width=4}
veh_group <- rnorm(10, 0, 1)
drug1_group <- rnorm(10, 1.5, 0.5)
drug2_group <- rnorm(10, 3, 4)
drug3_group <- rnorm(10, 3, 1.25)
effect_size_p_df <- data.frame(group=ofactor(rep(c("Vehicle", "Drug1", "Drug2", "Drug3"),
                                         each=10)),
                               value=c(veh_group, drug1_group, drug2_group, drug3_group))
effect_size_p_df %>%
  ggplot(aes(group, value, colour = group)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.09, height = 0, seed = 7)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "grey25") +
  tight_x() +
  scale_colour_manual(values = c("Vehicle"="darkgray",
                                 "Drug1"="lightblue",
                                 "Drug2"="orange",
                                 "Drug3"="#056f92"), guide = "none") +
  coord_cartesian(ylim = c(-3,10)) + 
  scale_y_continuous(oob=scales::squish) +
  labs(title = "Effect size and p-values",
       x = NULL, y = "Arbitrary unit")
```

```{r}
## I want to compare each group of drug to Vehicle using t-test, and get difference in mean values.
drug_test_lists <- split(effect_size_p_df$value, effect_size_p_df$group)
drug_t_tests <- do.call(rbind, lapply(drug_test_lists, function(x) {
  ttest <- t.test(x, drug_test_lists[[1]])
  estimates <- ttest$estimate
  pvalue <- ttest$p.value
  ci <- ttest$conf.int
  res <- c(diff=unname(estimates[1]-estimates[2]),
              p=pvalue,
              ci_low=ci[1],
              ci_high=ci[2])
})) %>% as.data.frame
drug_t_tests$Comparison <- paste0(rownames(drug_t_tests),
                                 " vs vehicle")
drug_t_tests <- dplyr::select(drug_t_tests,
                              "Comparison", "diff", "p", "ci_low", "ci_high")
gt::gt(drug_t_tests) %>%
  fmt_number(
    columns=c("diff", "ci_low", "ci_high"),
    decimals = 2,
    use_seps = FALSE
  ) %>%
  fmt_scientific(columns="p", decimals=2) %>%
  opt_row_striping(row_striping=TRUE)
```

```{r  effect_size_p_value_nonzero_diff, fig.height=4, fig.width=4}
effect_size_p_df %>%
  ggplot(aes(group, value, colour = group)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.09, height = 0, seed = 7)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "grey25") +
  tight_x() +
  geom_hline(yintercept=1.5, col="#B35", linetype=2, lwd=1.5) +
  scale_colour_manual(values = c("Vehicle"="darkgray",
                                 "Drug1"="lightblue",
                                 "Drug2"="orange",
                                 "Drug3"="#056f92"), guide = "none") +
  coord_cartesian(ylim = c(-3,10)) + 
  scale_y_continuous(oob=scales::squish) +
  labs(title = "Effect size and p-values",
  subtitle="We now consider a difference of 1.5 as meaningful",
       x = NULL, y = "Arbitrary unit")
```

```{r}
## I want to compare each group of drug to Vehicle using t-test, and get difference in mean values.
drug_t_tests_nonzero <- do.call(rbind, lapply(drug_test_lists, function(x) {
  ttest <- t.test(x, drug_test_lists[[1]], mu=1.5, alternative="greater")
  estimates <- ttest$estimate
  pvalue <- ttest$p.value
  ci <- ttest$conf.int
  res <- c(diff=unname(estimates[1]-estimates[2]),
              p=pvalue,
              ci_low=ci[1],
              ci_high=ci[2])
})) %>% as.data.frame
drug_t_tests_nonzero$Comparison <- paste0(rownames(drug_t_tests_nonzero),
                                 " vs vehicle")
drug_t_tests_nonzero$Threshold <- 1.5
drug_t_tests_nonzero <- dplyr::select(drug_t_tests_nonzero,
                              "Comparison", "Threshold", "diff", "p", "ci_low", "ci_high")
gt::gt(drug_t_tests_nonzero) %>%
  fmt_number(
    columns=c("diff", "ci_low", "ci_high"),
    decimals = 2,
    use_seps = FALSE
  ) %>%
  fmt_scientific(columns="p", decimals=2) %>%
  opt_row_striping(row_striping=TRUE)
```


### Confidence intervals, and what "95%" refers to

```{r m7-coverage, fig.width=3.5, fig.height=5}
set.seed(107)  ## fixed so the deck and handout stay in sync
coverage_sims <- replicate_studies(100, n_donors = 8, n_wells = 3, fold = 5)
## the simulated truth, known exactly - no need to estimate it from the runs
## (and an arithmetic mean of ratios would be the wrong summary anyway)
true_effect   <- thr$true_fold

coverage_sims %>%
  mutate(covers = lwr <= true_effect & upr >= true_effect,
         sim = rank(estimate, ties.method = "first")) %>%
  ggplot(aes(sim, estimate, colour = covers)) +
  geom_hline(yintercept = true_effect, linetype = "dashed", linewidth = 1) +
  geom_linerange(aes(ymin = lwr, ymax = upr), linewidth = 0.8) +
  geom_point(size = 1.1) +
  scale_y_log10() +
  coord_flip() +
  scale_colour_manual(values = c(`TRUE` = "grey55", `FALSE` = "#c1443b"),
                      labels = c(`TRUE` = "covers the ground truth",
                                 `FALSE` = "misses")) +
  labs(title = "Confindence intervals of 100 repeats",
       x = "Simulated study (sorted by estimated effect size)", y = "Estimated fold change") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        legend.position = "bottom")
```

"95% confidence" is a statement about the **procedure**, not about any one
interval: run the study forever and 95% of the intervals you build this way
will contain the truth. You never learn which kind you got. What the interval
does give you, immediately and usefully, is a range of effect sizes the data
are compatible with.

### Which result is more informative?

> **A.** p = 0.03
>
> **B.** estimated shift = 4.8-fold, plausible range 3.1 to 7.5-fold

Only B can be compared with a threshold. Only B supports a decision.

### Three results, three different conclusions

```{r m7-three, fig.width=6.6, fig.height=3}
three <- tribble(
  ~result,                     ~estimate, ~lwr, ~upr, ~p, 
  "A  large, very uncertain",      6.0,    1.2, 30.0, 0.15, 
  "B  tiny, very (or too) precise",         1.15,   1.08, 1.22, 1,
  "C  moderate, moderate CI",      4.8,    3.1,  7.5, 0.03) %>%
  mutate(verdict = case_when(
    lwr > MCID_FOLD  ~ "relevant difference",
    upr < MCID_FOLD  ~ "no relevant difference",
    TRUE        ~ "inconclusive"),
    verdict = factor(verdict, levels = c("relevant difference",
                                         "no relevant difference",
                                         "inconclusive")))

export_data(three, "m07_three_results",
            "three hypothetical results and what each licenses you to conclude")

ggplot(three, aes(estimate, fct_rev(result), colour = verdict)) +
  annotate("rect", xmin = 0, xmax = MCID_FOLD, ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_vline(xintercept = MCID_FOLD, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 1, colour = "grey60") +
  scale_x_log10() +
  geom_linerange(aes(xmin = lwr, xmax = upr), linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_discrete(expand = expansion(add = 0.42)) +
  geom_text(aes(label=paste0("p=", p)),
            vjust=1.5) +
  guides(colour = guide_legend(nrow = 2)) +
  scale_colour_manual(values = c("relevant difference" = "#3b7a57",
                                 "no relevant difference" = "#4c8fbd",
                                 "inconclusive" = "#d9a441")) +
  labs(title = "Interpreting confidence intervals",
       x = "Fold change in cytotoxicity IC50", y = NULL) +
  theme(legend.position = "bottom")
```

B is the interesting one: a precise, unambiguous, *statistically significant*
effect that is far too small to matter. Statistical significance and biological
importance are different questions, and only the second one was ever the
reason for running the assay.

### Three conclusions, not two

Most people are trained to report two outcomes. There are three:

1. **Evidence for a relevant difference** — the interval lies above the
   threshold
2. **Evidence against a relevant difference** — the interval lies entirely
   below it
3. **Inconclusive** — the interval spans it; the experiment cannot separate
   the two, and that is a result worth reporting

### Our study

```{r m7-verdict}
ft <- fold_test(folds)
verdict <- ft %>%
  transmute(fold_shift = estimate, ci_low = lwr, ci_high = upr,
            threshold = MCID_FOLD, p_value,
            conclusion = case_when(
              lwr > MCID_FOLD ~ "relevant difference - act",
              upr < MCID_FOLD ~ "no relevant difference - do not act",
              TRUE            ~ "inconclusive - more lots needed")) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

export_data(verdict, "m07_study_verdict",
            "the pre-registered decision applied to our data")
verdict
```

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

The p-value answers one narrow question about a world we do not believe in.
The effect size and its interval answer the question we asked. Report both,
decide on the second.

---

# Module VIII — What is power, and how does it affect the quality of our decision?

### Story

Back to the original question, with the whole dataset visible.

> Would you change the project's plan on the basis of these data? What
> experiment, if any, would you run next?

### Power, without formulas

The same true `r thr$true_fold`-fold difference, the same assay, the same
noise. Only the number of lots changes.

```{r m8-power, fig.width=7, fig.height=4.5}
set.seed(108)  ## fixed so the deck and handout stay in sync
power_grid <- map_dfr(c(2, 3, 4, 6, 12), function(n)
  replicate_studies(params$n_sim, n_donors = n, n_wells = 3,
                    fold = thr$true_fold))

power_summary <- power_grid %>%
  group_by(n_donors) %>%
  summarise(
    power_significant = mean(significant),
    power_decisive    = mean(calls_relevant),
    median_estimate   = median(estimate),
    median_ci_width   = median(upr - lwr),
    .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

export_data(power_summary, "m08_power_by_n",
            "power to reach significance vs power to clear the relevance threshold")
power_summary
```

```{r m8-power-fig, fig.width=7, fig.height=3}
p8a <- power_summary %>%
  pivot_longer(c(power_significant, power_decisive),
               names_to = "criterion", values_to = "power") %>%
  mutate(criterion = recode(criterion,
           power_significant = "p < 0.05",
           power_decisive    = "CI above the threshold")) %>%
  ggplot(aes(n_donors, power, colour = criterion)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 1.1) + geom_point(size = 2.6) +
  scale_x_continuous(breaks = c(2, 3, 4, 6, 12)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("#4c8fbd", "#3b7a57")) +
  labs(title = "Power increases with sample size",
       x = "Independent hepatocyte lots", y = "Proportion of studies") +
  theme(legend.position = "bottom")

p8b <- power_grid %>%
  mutate(n_lab = factor(paste0("n = ", n_donors),
                        levels = paste0("n = ", c(2, 3, 4, 6, 12)))) %>%
  ggplot(aes(estimate, n_lab, fill = n_lab)) +
  geom_violin(colour = NA, alpha = 0.75, width = 0.95) +
  geom_boxplot(width = 0.11, outlier.shape = NA, fill = "white") +
  scale_y_discrete(expand = expansion(add = 0.45)) +
  geom_vline(xintercept = median(power_grid$estimate[power_grid$n_donors == 24]),
             linetype = "dashed", linewidth = 1) +
  scale_x_log10() +
  geom_vline(xintercept=thr$true_fold, col="#333", linestyle= "dashed") +
  scale_fill_manual(values=royalbluegrayred(5), guide = "none") +
  coord_flip () +
  labs(title = "Distribution of estimated fold changes",
       x = "Estimated fold change", y = NULL)

p8b + p8a
```

Two things to take from panel A. First, power is not a property of an assay —
it is a property of an assay **plus a hypothesised effect size**. "How many
samples do I need?" has no answer until "what difference would matter?" has
one. Second, the two curves are far apart: with
`r power_summary$n_donors[1]` lots you would reach significance in
`r scales::percent(power_summary$power_significant[1])` of studies but could only
*act* on `r scales::percent(power_summary$power_decisive[1])` of them.

Panel B is the reason small studies are dangerous rather than merely weak. At
*n* = 3 the same experiment, run on the same compounds, can return almost any
answer.

There is a second, less obvious cost. A study that is underpowered can only
reach significance by landing on an unusually large estimate — so among the
studies that *do* get published, the effect is systematically overstated. The
5-fold difference above is too easy to detect for this to bite, so the
simulation below uses a genuinely marginal effect instead.

```{r m8-exaggeration, fig.width=7, fig.height=4.5}
set.seed(109)  ## fixed so the deck and handout stay in sync
marginal <- map_dfr(c(3, 4, 6, 12), function(n)
  replicate_studies(params$n_sim, n_donors = n, n_wells = 3, fold = 1.2))

## A fold shift is a ratio, so "how much is it exaggerated" is a question
## about the log. Comparing folds directly would understate it badly.
exaggeration <- marginal %>%
  group_by(n_donors) %>%
  summarise(power            = mean(significant),
            all_studies      = 10^median(log10(estimate)),
            significant_only = 10^median(log10(estimate[significant])),
            .groups = "drop") %>%
  mutate(exaggeration = log10(significant_only) / log10(all_studies),
         across(where(is.numeric), ~ round(.x, 2)))

export_data(exaggeration, "m08_exaggeration",
            "with a marginal true effect: median estimate among all studies vs among significant ones")
exaggeration
```

```{r m8-exaggeration-fig, fig.width=5.4, fig.height=3.5}
ggplot(exaggeration, aes(power, exaggeration)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 1.2, colour = "#c1443b") +
  geom_point(size = 2.8, colour = "#c1443b") +
  geom_text(aes(label = paste0("n = ", n_donors)), vjust = -1.2, size = GEOM_PT) +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0.95, 1.5)) +
  labs(title = "The less power a study has, the more it exaggerates",
       subtitle = "median effect among significant studies vs among all",
       x = "Power", y = "Exaggeration of the reported effect")
```

At `r scales::percent(exaggeration$power[1])` power the surviving studies
overstate the effect by about
`r scales::percent(exaggeration$exaggeration[1] - 1)`; by the time power
reaches `r scales::percent(exaggeration$power[nrow(exaggeration)])` the
distortion is gone. An underpowered literature is not merely a noisy one — it
is a biased one, and the bias always points the same way.

### Where to spend twenty-four wells

```{r m8-allocation-prep}
set.seed(110)  ## fixed so the deck and handout stay in sync
allocation <- bind_rows(
  replicate_studies(params$n_sim, n_donors = 2, n_wells = 12,
                    fold = thr$true_fold) %>%
    mutate(allocation = "2 donors × 12 replicates"),
  replicate_studies(params$n_sim, n_donors = 3, n_wells = 8,
                    fold = thr$true_fold) %>%
    mutate(allocation = "3 donors × 8 replicates"),
  replicate_studies(params$n_sim, n_donors = 8, n_wells = 3,
                    fold = thr$true_fold) %>%
    mutate(allocation = "8 donors × 3 replicates"),
  replicate_studies(params$n_sim, n_donors = 12, n_wells = 2,
                    fold = thr$true_fold) %>%
    mutate(allocation = "12 donors × 2 replicates"))  %>%
  mutate(allocation=ofactor(allocation))

allocation %>%
  group_by(allocation) %>%
  summarise(power_significant = mean(significant),
            power_decisive    = mean(calls_relevant),
            median_ci_width   = median(upr - lwr), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  export_data("m08_allocation",
              "24 wells spent in different ways")
```

```{r m8-allocation, fig.width=5, fig.height=5}
allocation %>%
  ggplot(aes(estimate, fill = allocation)) +
  geom_density(alpha = 0.55, colour = NA) +
  facet_wrap(~allocation, ncol=1) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05))) +
  geom_vline(xintercept = MCID_FOLD, linetype = "dashed", linewidth = 1) +
  scale_x_log10() +
  scale_fill_manual(values = royalbluegrayred(4)) +
  labs(title = "Balancing replicates to maximize power",
       x = "Estimated fold change", y = "Density")
```

Both use 24 wells per compound. Three lots with eight wells each measures three
lots very precisely and then has to generalize from three; eight lots with
three wells each measures each lot a little less precisely and generalizes from
eight. Since the uncertainty that matters is lot-to-lot, the second wins — and
it keeps winning no matter how many wells you add to the first.

### Over to you (5 min)

Think of one of your own experiments or assays where this idea applies. What
questions does it raise? How relevant is it to your work, and would it change
what you do?

### Takeaway

Power is not a number you compute at the end to justify what you did. It is
the question "what would this experiment be able to tell me?", asked while
there is still time to change the answer.

---

# What we have learned

One dataset, ten questions, and one answer that changed as we asked better
questions about the design:

```{r m9-journey}
journey <- bind_rows(
  tibble(step = "1. count wells, ignore lots (Module III)",
         fold_shift = pseudo$fold[1],
         ci_low = pseudo$ci_low[1], ci_high = pseudo$ci_high[1]),
  tibble(step = "2. confounded schedule (Module IV)",
         fold_shift = design_summary$fold_shift[
           design_summary$design == "Confounded"],
         ci_low = NA_real_, ci_high = NA_real_),
  tibble(step = "3. lots, but pairing ignored (Module V)",
         fold_shift = test_table$fold_shift[2],
         ci_low = test_table$ci_low[2], ci_high = test_table$ci_high[2]),
  tibble(step = "4. lots, paired, blocked (Modules V-VII)",
         fold_shift = test_table$fold_shift[3],
         ci_low = test_table$ci_low[3], ci_high = test_table$ci_high[3])) %>%
  mutate(threshold = MCID_FOLD,
         conclusion = case_when(
           is.na(ci_low)        ~ "no interval - the design cannot separate day from compound",
           ci_low > MCID_FOLD   ~ "relevant difference",
           ci_high < MCID_FOLD  ~ "no relevant difference",
           TRUE                 ~ "inconclusive"),
         earned = c("no - the interval is not entitled to its width",
                    "no - the estimate contains the run-day effect",
                    "yes",
                    "yes"),
         across(where(is.numeric), ~ round(.x, 2)))

export_data(journey, "m09_summary_of_analyses",
            "the same 48 numbers analyzed four ways")
journey
```

Same 48 numbers every time. Only step 4 is both right and *entitled* to be
right.

Step 1 is the instructive one. Counting wells instead of lots produced the
smallest p-value on the page — `r pseudo$p_value[1]` — and still reached the
wrong verdict, because the same analysis also ignored the pairing. A very small
p-value is not evidence that the analysis was the right one. Here it is
evidence of the opposite.

Step 2 has no interval at all. That is not a missing number; it is the design
telling you that the question cannot be answered from these data.

### The three things worth remembering

* Good design **minimizes unnecessary and avoidable variation** while
  maximizing the chance of detecting the effect you care about. It decides
  what the statistics will be able to separate later.
* Design and analysis **reduce uncertainty; they never remove it.** An honest
  interval is a better deliverable than a confident sentence.
* The decision is **scientific**. Effect size, threshold, context and prior
  knowledge live outside statistics; the analysis only tells you how much of
  the data's uncertainty you have to carry into that decision.

### The one-line version

> To answer *"which statistical test should I use?"* you first have to answer
> *"how were the data generated?"*

---

# Module IX — How can I learn more and get help if I am stuck?

### Getting help from an LLM

This is where the workshop pays off in practice. Consider the usual prompt:

> *"Compare columns A and B statistically."*

An LLM will answer it. It will pick a test, report a p-value, and sound
confident. It cannot possibly know whether the columns are paired, whether the
rows are wells or lots, whether the two groups were run on different days, or
what difference you would act on. So it will guess, and the guess is usually
an unpaired t-test on every row — which, on this dataset, is exactly the
analysis that gets the wrong answer twice over.

Now the same request with the design in it:

> *"These measurements come from eight independent human hepatocyte lots. Each
> lot was tested with both compounds, with three technical replicate wells per
> lot and compound. The independent experimental unit is the lot, not the
> well. The plates were run over two days; day is balanced across compounds.
> We decided in advance that a shift in the cytotoxicity IC50 smaller than
> 3-fold would not change our decision. Help me visualize the data so the
> pairing is visible, estimate the compound difference with its uncertainty,
> and compare that interval with our threshold. Before you recommend a test,
> tell me what you think the independent unit is and whether the comparison is
> paired."*

Everything that makes the second prompt work is a design fact, and every one
of those facts was decided before any data existed. The final sentence is the
most useful habit: **ask the model to state its assumptions about the unit and
the pairing before it computes anything**, so that you can catch a wrong
assumption while it is still cheap.

The same applies to a colleague, a statistician, or your own notebook six
months from now.

### Data to take away

Every module exported the data it used. They are in
`r params$outdir`, as plain CSV — open them in Excel, Prism, R, Python, or
paste one into an LLM along with the design description above.

```{r m9-manifest}
.manifest %>% arrange(file)
```

```{r m9-full-exports}
## the complete dataset in both shapes, plus the hidden truth
export_data(viab, "full_dataset_long",
            "all 48 wells with plate references (long format)")

export_data(
  viab %>%
    select(donor_id, donor_sex, donor_age, weekday, compound, replicate,
           viability_pct) %>%
    pivot_wider(names_from = c(compound, replicate),
                values_from = viability_pct,
                names_glue = "{compound}_well{replicate}"),
  "full_dataset_wide",
  "all 48 wells, one row per lot (wide format, for Excel/Prism)")

export_data(
  sim$truth %>%
    select(donor_id, compound, ic50_true, viab_true, sensitive) %>%
    mutate(across(where(is.numeric), ~ round(.x, 2))),
  "answer_key_true_values",
  "the simulated ground truth - the answer key, for after the workshop")
```

### Reading

* *Points of Significance* series of Nature Methods
* Lazic, *Experimental Design for Laboratory Biologists* (2016)

### And otherwise

Come and talk to me. Most statistical problems that reach a statistician are
design problems that arrived too late, and the conversation is much shorter
and much happier when it happens **before** the experiment.

---

# Appendix — the companion document

`2026-09-statistical-thinking.Rmd` takes the same assay to full analyst depth
and is the place to look when a module raised a question this notebook
deliberately simplified:

| Topic | Where |
|---|---|
| Full 96-well plate layout, edge effects | *Design* |
| Eight concentrations, four-parameter logistic fits | *Curve fitting* |
| Absolute vs relative IC~50~, and why the difference matters | *Curve fitting* |
| Pre-specified plate QC: Z', vehicle CV, signal window | *Plate quality control* |
| `alias()` and formally undiagnosable designs | *Consequences of the two designs* |
| A whole plate failing technically | *Perturbation 1* |
| Grubbs vs a genuinely sensitive lot | *Perturbation 2* |
| Weighted fitting, biological vs technical variance | *Perturbation 3* |

# Session information

```{r session}
sessionInfo()
```
