---
title: "Statistical thinking for experimental scientists"
subtitle: "Ten modules around one small CellTiter-Glo dataset"
author: "Jitao David Zhang"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    number_sections: false
    code_folding: hide
    fig_width: 7
    fig_height: 4.5
params:
  seed: 20260919
  n_sim: 2000          # Monte Carlo replicates for modules 6 and 8
  outdir: "output/workshop"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "png", dpi = 200,
                      ## figures are drawn at their final size: 7 inches or
                      ## less, so 12 pt text really is 12 pt on the page
                      fig.width = 7, fig.height = 4.5,
                      ## keep every figure on disk: the slide deck and the
                      ## one-pager are built from these files by the Makefile
                      fig.path = "output/figures/")
```

```{r libraries}
library(tidyverse)
library(patchwork)
library(ribiosUtils)   # ofactor, percentage
library(ribiosPlot)    # openFileDevice, closeFileDevice

source("R/simulate_workshop_data.R")
set.seed(params$seed)
dir.create(params$outdir, showWarnings = FALSE, recursive = TRUE)

## One theme for the whole document. Nothing is below 12 pt, and every figure
## is drawn at 7 inches wide or less, so the text keeps a generous share of
## the canvas instead of shrinking when the figure is placed on a slide.
BASE_PT <- 12                  # smallest type anywhere in a figure
GEOM_PT <- BASE_PT / .pt       # the same size for geom_text()/annotate(), mm

theme_ws <- theme_bw(base_size = BASE_PT) +
  theme(plot.title    = element_text(face = "bold", size = BASE_PT + 1.5),
        plot.subtitle = element_text(size = BASE_PT, colour = "grey30"),
        axis.title    = element_text(size = BASE_PT),
        axis.text     = element_text(size = BASE_PT),
        strip.text    = element_text(size = BASE_PT, face = "bold"),
        legend.text   = element_text(size = BASE_PT),
        legend.title  = element_blank(),
        plot.title.position = "plot",
        legend.key.size = unit(0.9, "lines"),
        plot.margin   = margin(4, 6, 4, 4),
        panel.grid.minor = element_blank())
theme_set(theme_ws)

cpd_cols <- c("CPD-A" = "#1f6fb4", "CPD-B" = "#c1443b")
day_cols <- c("Friday" = "#e07b39", "Monday" = "#4c8fbd")
ok_cols  <- c("yes" = "#3b7a57", "no" = "#a33", "maybe" = "#d9a441")

## every module exports the data it uses, so participants can reopen it
## afterwards in Excel, Prism, or with an LLM
.manifest <- tibble(file = character(), rows = integer(), what = character())

## keep exported numbers readable without corrupting raw counts or p-values
tidy_num <- function(x) {
  if (!is.double(x)) return(x)
  nz <- abs(x[is.finite(x) & x != 0])
  if (length(nz) && min(nz) < 1e-3) signif(x, 3) else round(x, 3)
}

export_data <- function(x, name, what) {
  path <- file.path(params$outdir, paste0(name, ".csv"))
  write_csv(mutate(x, across(where(is.double), tidy_num)), path)
  .manifest <<- add_row(.manifest, file = basename(path),
                        rows = nrow(x), what = what)
  invisible(x)
}
```

```{r the-data}
sim   <- simulate_workshop_data(seed = params$seed)
viab  <- sim$viab                 # 48 wells, the dataset the workshop uses
dm    <- donor_means(viab)        # 16 lot-level means, the analysis unit
thr   <- sim$thresholds
MCID  <- thr$mcid_pp              # relevance threshold, in percentage points
CONC  <- sim$params$screen_conc
```

# How to use this notebook

Every module follows the same five-step loop, and the loop is the point of
the workshop:

> **Question → Design → Data → Analysis and uncertainty → Decision**

One dataset carries all ten modules. It is synthetic, so the true answer is
known and every claim in here can be checked against it.

**The experiment.** Two compounds from the same project, CPD-A and CPD-B, are
tested for cytotoxicity in primary human hepatocytes with the CellTiter-Glo
ATP readout.

| | |
|---|---|
| Independent units | 8 hepatocyte lots (donors) |
| Conditions | CPD-A and CPD-B, both in every lot |
| Technical replicates | 3 wells per lot × compound |
| Readout concentration | `r CONC` µM, one point on the curve |
| Run days | Friday and Monday |
| Total measurements | 48 |

Three things are hidden in it, and each surfaces in the module where it
matters: a run-day effect that lives in the vehicle reference, one well with
an air bubble, and one lot that is genuinely more sensitive than the rest.

> **A note on the numbers.** This notebook is the participants' document. A
> companion document, `2026-09-statistical-thinking.Rmd`, works the same assay
> at full analyst depth: 96-well plates, eight concentrations, four-parameter
> logistic fits, plate QC, weighted regression and variance decomposition.
> Modules 4 and 7 point into it. The two documents use the same compounds and
> the same day effect but are simulated separately, so the individual numbers
> differ.

---

# Module 1 — What question are we actually asking?

### Story

A project team wants to know whether to drop CPD-B. Someone says "run the
viability assay and tell me if it's significant". That instruction cannot be
carried out, because it does not say what would count as an answer.

Five questions turn it into an experiment:

* Measurements **of what**? — ATP content, as a proxy for viable cells
* From **whom or what**? — primary human hepatocytes, one lot per donor
* Under **which conditions**? — `r CONC` µM, 24 h, 0.1% DMSO vehicle
* What **population** do we want to speak about? — human hepatocyte lots in
  general, not these eight
* What **decision** does this inform? — whether CPD-B needs a dedicated
  safety study

### The sentence every experiment should be able to complete

> I want to compare **viability at `r CONC` µM** in **primary human
> hepatocyte lots** because I need to decide **whether CPD-B is materially
> less cytotoxic than CPD-A**.

### The threshold comes before the data

The team agrees, before seeing anything, that a **`r thr$mcid_fold`-fold shift
in potency** is what they would act on. That is a scientific judgement, not a
statistical one — no analysis can supply it.

The assay does not report potency, it reports viability at one concentration.
So the threshold has to be translated. The translation is not a single number,
because the same potency shift produces a different viability gap depending on
where a lot sits on the curve.

```{r m1-threshold, fig.width=7, fig.height=8.6}
curve_df <- expand_grid(
    series = c("CPD-A reference", "a 3-fold less potent compound"),
    conc   = 10^seq(log10(0.3), log10(1000), length.out = 300)) %>%
  mutate(ic50 = if_else(series == "CPD-A reference",
                        sim$params$ic50_A,
                        sim$params$ic50_A * thr$mcid_fold),
         viability = 100 * hill_viability(conc, ic50, sim$params$hill_A))

v_lo <- 100 * hill_viability(CONC, sim$params$ic50_A, sim$params$hill_A)
v_hi <- 100 * hill_viability(CONC, sim$params$ic50_A * thr$mcid_fold,
                             sim$params$hill_A)

p1a <- ggplot(curve_df, aes(conc, viability, colour = series)) +
  geom_line(linewidth = 1.3) +
  annotate("segment", x = CONC, xend = CONC, y = v_lo, yend = v_hi,
           linewidth = 1.2, colour = "grey25",
           arrow = arrow(ends = "both", length = unit(0.18, "cm"))) +
  annotate("label", x = CONC, y = (v_lo + v_hi) / 2,
           label = sprintf("%.0f pp", v_hi - v_lo), size = GEOM_PT) +
  geom_vline(xintercept = CONC, linetype = "dotted", colour = "grey40") +
  scale_x_log10() +
  scale_colour_manual(values = c("CPD-A reference" = "#1f6fb4",
                                 "a 3-fold less potent compound" = "#7a3b8f")) +
  labs(title = "A  A 3-fold potency shift, seen at one concentration",
       subtitle = sprintf("read out at %.1f µM", CONC),
       x = "Concentration (µM)", y = "ATP (% of vehicle)") +
  theme(legend.position = "bottom")

p1b <- sim$mcid_by_donor %>%
  ggplot(aes(fct_reorder(donor_id, mcid_pp), mcid_pp)) +
  geom_col(fill = "#7a3b8f", width = 0.7) +
  geom_hline(yintercept = MCID, linetype = "dashed", linewidth = 1) +
  geom_hline(yintercept = thr$mcid_pp_naive, linetype = "dotted",
             colour = "grey35", linewidth = 1) +
  annotate("text", x = 0.6, y = MCID + 2.2, hjust = 0, size = GEOM_PT,
           label = sprintf("average: %.1f pp", MCID)) +
  annotate("text", x = 0.6, y = thr$mcid_pp_naive + 2.2, hjust = 0,
           size = GEOM_PT, colour = "grey35",
           label = sprintf("one 'typical' lot: %.1f pp", thr$mcid_pp_naive)) +
  expand_limits(y = 45) +
  labs(title = "B  The same threshold, lot by lot",
       subtitle = "a lot at the top or bottom of the curve shows a smaller gap",
       x = NULL, y = "Gap equal to a 3-fold shift (pp)")

p1a / p1b
```

So the pre-registered rule becomes: **we will act if the average viability
difference between the compounds exceeds `r round(MCID, 1)` percentage
points.** Note that translating at a single "typical" lot would have given
`r round(thr$mcid_pp_naive, 1)` pp — over half again as large, and it would
have made the study look like a failure. A threshold has to be expressed on
the same scale, and in the same units, as the estimate it will be compared
with.

```{r m1-export}
export_data(sim$mcid_by_donor, "m01_relevance_threshold",
            "the 3-fold rule translated into a viability gap, lot by lot")
```

### Your turn (5 min)

Complete the sentence for one of your own experiments:

> I want to estimate/compare ______ in ______ because I need to decide ______.

Then name the number that would make you act. If you cannot, the experiment
is not ready to be designed.

### Takeaway

Statistics cannot rescue an experiment whose question was never defined. The
threshold is a scientific decision, it is fixed before the data, and it must
be carried onto the scale you will actually measure.

---

# Module 2 — Variation: what does one number tell us?

### Story

The first lot comes off the plate reader. Six numbers. What can be said?

```{r m2-reveal}
one_lot <- viab %>%
  filter(donor_id == "HH1") %>%
  select(donor_id, compound, replicate, raw_rlu, viability_pct)

export_data(one_lot, "m02_one_lot",
            "six wells from a single hepatocyte lot, both compounds")
one_lot
```

### Show the points, not a bar

```{r m2-dots, fig.width=7, fig.height=3.9}
p2a <- one_lot %>%
  ggplot(aes(compound, viability_pct, colour = compound)) +
  stat_summary(fun = mean, geom = "col", fill = "grey85", colour = NA,
               width = 0.6) +
  stat_summary(fun.data = mean_se, geom = "errorbar", width = 0.2,
               colour = "grey30") +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  labs(title = "What a bar chart shows", subtitle = "two numbers and an error bar",
       x = NULL, y = "Viability (% of vehicle)") +
  coord_cartesian(ylim = c(0, 80))

p2b <- one_lot %>%
  ggplot(aes(compound, viability_pct, colour = compound)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.08, height = 0, seed = 1)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45,
               colour = "grey25", linewidth = 0.6) +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  labs(title = "What the data show", subtitle = "three wells per compound, and their mean",
       x = NULL, y = NULL) +
  coord_cartesian(ylim = c(0, 80))

p2a | p2b
```

The bar chart is not wrong. It is simply compatible with almost any underlying
data, which is why it is a poor default.

### The mean moves every time

Nothing about the lot or the compound changes below. The same measurement
process is repeated twenty times, three wells at a time.

```{r m2-mean-moves, fig.width=7, fig.height=4.2}
set.seed(101)  ## fixed so the deck and handout stay in sync
true_v <- 100 * hill_viability(CONC,
                               sim$truth$ic50_true[sim$truth$donor_id == "HH1" &
                                                   sim$truth$compound == "CPD-B"],
                               sim$params$hill_A)
tech_cv <- 0.05 + 0.12 * (1 - true_v / 100)

repeats <- tibble(experiment = rep(1:20, each = 3)) %>%
  mutate(well = rep(1:3, 20),
         viability = true_v * (1 + rnorm(n(), 0, tech_cv)))

rep_means <- repeats %>%
  group_by(experiment) %>%
  summarise(mean_viability = mean(viability), .groups = "drop")

ggplot(repeats, aes(experiment, viability)) +
  geom_hline(yintercept = true_v, linetype = "dashed", colour = "#c1443b",
             linewidth = 1) +
  geom_point(colour = "grey55", size = 1.9, alpha = 0.8) +
  geom_point(data = rep_means, aes(y = mean_viability), size = 2.9,
             colour = "#1f6fb4") +
  geom_line(data = rep_means, aes(y = mean_viability), colour = "#1f6fb4",
            linewidth = 0.9) +
  annotate("text", x = 20.4, y = true_v, hjust = 1, vjust = -0.8, size = GEOM_PT,
           colour = "#c1443b", label = "the unchanging truth") +
  labs(title = "Twenty repeats of one unchanged experiment",
       subtitle = "grey: individual wells   ·   blue: the mean of each repeat",
       x = "Repeat of the experiment", y = "Viability (% of vehicle)")
```

The truth is a fixed line. Every mean sits somewhere else. That gap is
**sampling uncertainty**, and it exists even when the experiment is perfect.
A confidence interval, later, is nothing more than an attempt to say how wide
this cloud is — from a single experiment.

### Not all variation is the same

```{r m2-taxonomy, fig.width=7, fig.height=4.4}
set.seed(102)  ## fixed so the deck and handout stay in sync
n_lots <- 8
base   <- 100 * hill_viability(CONC, sim$params$ic50_A, sim$params$hill_A)
lot_shift <- 100 * hill_viability(
  CONC, sim$params$ic50_A * 10^rnorm(n_lots, 0, log10(sim$params$donor_gsd)),
  sim$params$hill_A)

taxonomy <- bind_rows(
  tibble(layer = "1. measurement only",
         value = base * (1 + rnorm(n_lots * 3, 0, 0.06))),
  tibble(layer = "2. + lot biology",
         value = rep(lot_shift, each = 3) * (1 + rnorm(n_lots * 3, 0, 0.06))),
  tibble(layer = "3. + a run-day shift",
         value = rep(lot_shift, each = 3) * (1 + rnorm(n_lots * 3, 0, 0.06)) *
                 rep(c(1, 1 / 1.2), each = n_lots * 3 / 2))
) %>%
  mutate(layer = ofactor(layer))

ggplot(taxonomy, aes(layer, value, colour = layer)) +
  geom_point(position = position_jitter(width = 0.14, seed = 3),
             size = 2.2, alpha = 0.8) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.5, colour = "grey25") +
  scale_colour_brewer(palette = "Dark2", guide = "none") +
  labs(title = "The same 24 wells, with variance sources switched on one at a time",
       subtitle = "spread is not one thing, and the three layers are not equally fixable",
       x = NULL, y = "Viability (% of vehicle)")
```

Four kinds, and they call for four different responses:

| Kind | Example here | What to do |
|---|---|---|
| **Wanted** | the compound effect | measure it |
| **Unavoidable** | well-to-well pipetting | replicate, then average |
| **Controllable** | run day, plate, operator | block it, or standardise it |
| **Confounding** | run day that tracks compound | redesign — it cannot be fixed later |

### Bias and variance

```{r m2-target, fig.width=7, fig.height=7}
set.seed(103)  ## fixed so the deck and handout stay in sync
target_data <- expand_grid(
    bias     = c("unbiased", "biased"),
    variance = c("low variance", "high variance")) %>%
  mutate(shots = pmap(list(bias, variance), function(b, v) {
    ctr <- if (b == "biased") c(1.45, 1.05) else c(0, 0)
    s   <- if (v == "high variance") 1.15 else 0.32
    tibble(x = rnorm(24, ctr[1], s), y = rnorm(24, ctr[2], s))
  })) %>%
  unnest(shots) %>%
  mutate(bias     = factor(bias, levels = c("unbiased", "biased")),
         variance = factor(variance, levels = c("low variance", "high variance")))

## target rings, drawn as polygons so no extra package is needed
rings <- expand_grid(r = c(0.6, 1.2, 1.8, 2.4),
                     theta = seq(0, 2 * pi, length.out = 120)) %>%
  mutate(x = r * cos(theta), y = r * sin(theta))

ggplot(target_data) +
  geom_path(data = rings, aes(x, y, group = r), colour = "grey75") +
  geom_point(aes(x, y), colour = "#c1443b", size = 1.7, alpha = 0.85) +
  geom_point(aes(x = 0, y = 0), colour = "grey40", size = 1.6, shape = 3) +
  facet_grid(variance ~ bias) +
  coord_equal(xlim = c(-2.6, 2.6), ylim = c(-2.6, 2.6)) +
  labs(title = "Bias is consistently wrong. Variance is inconsistently noisy.",
       subtitle = "the cross is the truth; reducing variance is only a gain if it does not buy bias",
       x = NULL, y = NULL) +
  theme(axis.text = element_blank(), axis.ticks = element_blank(),
        panel.grid = element_blank())
```

Two scenarios to argue about:

* **Scenario A.** Every control sample is pipetted with pipette A, every
  treated sample with pipette B.
* **Scenario B.** Within one experiment, every replicate uses a different
  consumable batch, "to make it representative".

> Which one makes it hard to know whether the treatment caused the difference?
> Which one makes a real effect harder to detect?

### Your turn (5 min)

Name one source of variability in your own assay, and classify it: wanted,
unavoidable, controllable, or confounding.

### Takeaway

Variation is not a sign of a bad experiment. Some of it is the signal, some is
the price of measuring, some is a design choice you still have, and some is a
design mistake you no longer have.

---

# Module 3 — What is *n*?

### Dataset reveal

All 48 measurements are now on the table.

```{r m3-reveal}
export_data(viab %>% select(donor_id, donor_sex, donor_age, compound, weekday,
                            plate_id, well, replicate, conc_uM, raw_rlu,
                            viability_pct),
            "m03_all_wells_long",
            "the full 48-well dataset, one row per well (long format)")

viab %>%
  select(donor_id, compound, weekday, replicate, viability_pct) %>%
  head(12)
```

> We have 48 numbers. What is *n*? 48? 16? 8? Something else?

### Two kinds of replicate

**Technical replicate** — a repeated measurement made to characterise the
*measurement* process. Three wells from one lot tell you how reproducibly you
can pipette and read that lot.

**Biological (independent) replicate** — an independently sampled unit that
carries new information about the *population*. A second hepatocyte lot is a
second donor, a second liver, a second draw from the population you want to
speak about.

Three wells from one lot are not three lots. Averaging the three wells gives
you a better measurement of **that one lot**; it does not give you a second
lot.

Here, the independent experimental unit is the **hepatocyte lot**: *n* = 8.

### Pseudoreplication, made visible

The same comparison, run twice. Nothing changes except what is treated as
independent.

```{r m3-pseudoreplication}
wrong <- t.test(viability_pct ~ compound, data = viab)              # 48 wells
right <- t.test(viability ~ compound, data = dm)                    # 8 lots

## both tests are CPD-A minus CPD-B; flip everything to CPD-B minus CPD-A
pseudo <- tibble(
  analysis = c("wells treated as independent (n = 48)",
               "lots treated as independent (n = 8)"),
  n_per_group = c(24, 8),
  estimate_pp = -c(diff(rev(wrong$estimate)), diff(rev(right$estimate))),
  ci_low  = c(-wrong$conf.int[2], -right$conf.int[2]),
  ci_high = c(-wrong$conf.int[1], -right$conf.int[1]),
  p_value = c(wrong$p.value, right$p.value)) %>%
  mutate(ci_width = ci_high - ci_low,
         across(c(estimate_pp, ci_low, ci_high, ci_width), ~ round(.x, 1)),
         p_value = signif(p_value, 3))

export_data(pseudo, "m03_pseudoreplication",
            "the same comparison analysed at the well level and at the lot level")
pseudo
```

The point estimate is identical to the first decimal. The p-value is about
`r round(pseudo$p_value[2] / pseudo$p_value[1])` times **smaller** when wells
are counted as independent, and the interval is about half as wide. Counting
wells adds no information about lots; it only makes the software believe there
is more of it. The extra confidence is manufactured.

One honest caveat, visible in the table. The pseudoreplicated interval here is
not just narrower — it happens to sit close to the answer we will reach in
Module 5 by doing things properly. That is luck, not vindication: ignoring the
lot structure double-counts wells, which narrows the interval, while ignoring
the pairing widens it, and on this dataset the two errors partly cancel. An
analysis that is wrong for two reasons at once can still land in a plausible
place. What it cannot do is *tell you* that it has.

### Twelve measurements, two ways to spend them

```{r m3-designs, fig.width=7, fig.height=4.6}
set.seed(104)  ## fixed so the deck and handout stay in sync
gp <- sim$params

## Experiment A: 12 lots x 1 well   |   Experiment B: 3 lots x 4 wells
one_study <- function(n_lots, n_wells) {
  ic50 <- gp$ic50_A * 10^rnorm(n_lots, 0, log10(gp$donor_gsd))
  v    <- 100 * hill_viability(gp$screen_conc, ic50, gp$hill_A)
  cv   <- 0.05 + 0.12 * (1 - v / 100)
  obs  <- v * (1 + matrix(rnorm(n_lots * n_wells), nrow = n_lots) * cv)
  lot_means <- rowMeans(obs)
  tibble(
    ## how well can we see how much lots differ from each other?
    sd_between_lots = if (n_lots > 1) sd(lot_means) else NA_real_,
    ## how well can we see how reproducible the measurement is?
    sd_within_lot   = if (n_wells > 1) mean(apply(obs, 1, sd)) else NA_real_)
}

design_cmp <- bind_rows(
  map_dfr(1:params$n_sim, ~ one_study(12, 1)) %>%
    mutate(design = "A: 12 lots\n× 1 well"),
  map_dfr(1:params$n_sim, ~ one_study(3, 4)) %>%
    mutate(design = "B: 3 lots\n× 4 wells"))

export_data(design_cmp %>% group_by(design) %>%
              summarise(across(everything(),
                               list(median = ~ median(.x, na.rm = TRUE),
                                    iqr = ~ IQR(.x, na.rm = TRUE)))),
            "m03_design_comparison",
            "12 measurements spent two ways: what each design can estimate")

design_cmp %>%
  pivot_longer(starts_with("sd_"), names_to = "quantity", values_to = "estimate") %>%
  mutate(quantity = recode(quantity,
           sd_between_lots = "spread BETWEEN lots\n(biological)",
           sd_within_lot   = "spread WITHIN a lot\n(technical)")) %>%
  ggplot(aes(design, estimate, fill = design)) +
  geom_violin(alpha = 0.6, colour = NA, na.rm = TRUE) +
  geom_boxplot(width = 0.14, outlier.shape = NA, na.rm = TRUE) +
  facet_wrap(~ quantity, scales = "free_y") +
  scale_fill_manual(values = c("#1f6fb4", "#c1443b"), guide = "none") +
  labs(title = "Both designs use 12 measurements",
       subtitle = "empty panel = that design cannot estimate that quantity at all",
       x = NULL, y = "Estimated SD (pp)")
```

Twelve lots × one well estimates lot-to-lot variation well and technical
variation not at all. Three lots × four wells does the opposite, and estimates
lot-to-lot spread from three points — which is barely an estimate. Neither is
better; they answer different questions. If the question is about *hepatocyte
lots in general*, spend the measurements on lots.

### Your turn (5 min)

In your last experiment: what was the independent unit? Was the number you
reported as *n* the number of units, or the number of measurements?

### Takeaway

Variation among biological replicates tells you about the population.
Variation among technical replicates tells you about your hands and your
instrument. *n* is the number of independent units, and the analysis has to
be done at that level.

---

# Module 4 — Design: blocking, randomisation, standardisation

### Story

Sixteen plates cannot be run in one sitting, so the assay is split over Friday
and Monday. Suppose the scheduler had done the natural thing and run **all
CPD-A plates on Friday and all CPD-B plates on Monday**.

Something real happened on Friday: the vehicle wells read about
`r scales::percent(sim$params$friday_vehicle_gain)` higher. Because viability is
expressed relative to each plate's own vehicle, a shift in the reference
propagates into every normalised value on that plate.

```{r m4-vehicle, fig.width=7, fig.height=4}
vehicle_wells <- sim$ctg %>%
  filter(design == "Blocked", well_type == "vehicle") %>%
  select(donor_id, compound, weekday, replicate, raw_rlu)

export_data(vehicle_wells, "m04_vehicle_wells",
            "vehicle (untreated) wells only - where the run-day effect lives")

vehicle_wells %>%
  ggplot(aes(weekday, raw_rlu / 1e6, fill = weekday)) +
  geom_boxplot(width = 0.45, alpha = 0.65, outlier.shape = NA) +
  geom_point(position = position_jitter(width = 0.1, seed = 2), size = 1.8,
             alpha = 0.7) +
  scale_fill_manual(values = day_cols, guide = "none") +
  labs(title = "The run-day effect is visible in the vehicle wells alone",
       subtitle = "no compound data needed - this diagnostic works before any comparison is made",
       x = NULL, y = expression("Raw signal (10"^6~"RLU)"))
```

### The same lots, two schedules

```{r m4-designs, fig.width=7, fig.height=4.6}
sim$schedule %>%
  select(donor_id, compound, Blocked = weekday_blocked,
         Confounded = weekday_confounded) %>%
  pivot_longer(c(Blocked, Confounded), names_to = "design",
               values_to = "weekday") %>%
  export_data("m04_schedules", "run-day allocation under the two designs") %>%
  ggplot(aes(compound, fct_rev(donor_id), fill = weekday)) +
  geom_tile(colour = "white", linewidth = 1.4) +
  geom_text(aes(label = str_sub(weekday, 1, 3)), colour = "white",
            fontface = "bold", size = GEOM_PT) +
  facet_wrap(~ design) +
  scale_fill_manual(values = day_cols) +
  labs(title = "The only difference between the designs is which day each plate ran",
       subtitle = "left: day varies within compound (a block)   ·   right: day tracks compound (a confounder)",
       x = NULL, y = "Hepatocyte lot") +
  theme(panel.grid = element_blank(), legend.position = "bottom")
```

### What each design concludes

```{r m4-consequences}
dm_conf <- donor_means(sim$viab_conf)

pair_up <- function(d) {
  d %>% select(donor_id, compound, viability) %>%
    pivot_wider(names_from = compound, values_from = viability) %>%
    mutate(difference = `CPD-B` - `CPD-A`)
}

design_result <- bind_rows(pair_up(dm)      %>% mutate(design = "Blocked"),
                           pair_up(dm_conf) %>% mutate(design = "Confounded"))

## the difference we would see with no day effect and no measurement noise
true_difference <- sim$truth %>%
  select(donor_id, compound, viab_true) %>%
  pivot_wider(names_from = compound, values_from = viab_true) %>%
  summarise(x = mean(`CPD-B` - `CPD-A`)) %>% pull(x)

design_summary <- design_result %>%
  group_by(design) %>%
  summarise(mean_difference_pp = mean(difference), .groups = "drop") %>%
  mutate(simulated_truth_pp = true_difference,
         error_pp           = mean_difference_pp - true_difference,
         across(where(is.numeric), ~ round(.x, 1)))

export_data(design_result, "m04_blocked_vs_confounded",
            "per-lot compound difference under both schedules")
design_summary
```

Both schedules land within a few percentage points of the truth, and on this
dataset the confounded one is not obviously worse. That is the difficulty with
confounding: it does not announce itself.

### The experiment that makes it obvious

So run the experiment that settles it. Simulate the same study with the two
compounds made **exactly identical** — same potency, same slope — and leave
everything else alone, including the Friday vehicle shift. Any difference the
analysis now reports is, by construction, not a compound difference.

```{r m4-null-compounds}
null_cpd <- simulate_workshop_data(seed = params$seed,
                                   ic50_B = sim$params$ic50_A)

test_pair <- function(d, label) {
  w  <- pair_up(donor_means(d))
  tt <- t.test(w$`CPD-B`, w$`CPD-A`, paired = TRUE)
  tibble(schedule = label, estimate_pp = tt$estimate,
         ci_low = tt$conf.int[1], ci_high = tt$conf.int[2],
         p_value = tt$p.value)
}

null_result <- bind_rows(
  test_pair(null_cpd$viab,      "Blocked"),
  test_pair(null_cpd$viab_conf, "Confounded")) %>%
  mutate(verdict = if_else(p_value < 0.05,
                           "a 'significant' difference between identical compounds",
                           "correctly finds nothing"),
         across(c(estimate_pp, ci_low, ci_high), ~ round(.x, 1)),
         p_value = signif(p_value, 3))

export_data(null_result, "m04_identical_compounds",
            "two IDENTICAL compounds run under both schedules")
null_result
```

The blocked schedule reports nothing, correctly. The confounded schedule
reports a difference of
`r null_result$estimate_pp[2]` pp with *p* =
`r null_result$p_value[2]` — statistically significant, perfectly
reproducible, and entirely an artefact of which day the plates ran.

Nothing in the confounded dataset can reveal this. Day and compound are the
same column: every CPD-A plate is a Friday plate. No test, no model and no
amount of care in the analysis can separate two factors that never varied
independently. In the blocked schedule day varies *within* each compound, so
it becomes a nuisance you can estimate rather than one you cannot see.

> One subtlety worth naming, because it shows up in the table above. The day
> effect is **multiplicative** (the vehicle reference moves by a factor) while
> we are reporting an **additive** difference in percentage points. Blocking
> makes the day effect cancel exactly in a *ratio*, but only approximately in
> a difference. Choosing a scale on which your effect is additive is the same
> decision we made in Module 1, and it keeps repeating.

### The three levers

| Lever | Use it for | Effect | Failure mode |
|---|---|---|---|
| **Block** | known, structural sources: day, plate, operator | removes confounding | blocking on something that tracks treatment does nothing |
| **Randomise** | everything you cannot block or foresee: well position, order | converts unknown bias into honest variance | small experiments can still randomise into an imbalance |
| **Standardise** | irrelevant sources: same pipette, one consumable lot | shrinks variance | standardising *per arm* manufactures bias |

The last column is where the damage is done. Using the same pipette throughout
reduces variance. Using pipette A for condition A and pipette B for condition B
is *also* "standardising", and it creates a confounder. Deliberately varying
consumable batch within every small experiment "to be representative" is not
robustness — it is added noise that makes a real effect harder to see.

### Your turn — the design poker (5 min)

In teams of two or three. Each team has four lots (A, 2, 3, 4), two compounds
(two colours) and two protocols (two shapes). Lay the experiments out in two
rows: Friday and Monday.

* Which factors did you block?
* Which did you randomise?
* Is any factor now perfectly aliased with another? If so, redeal.

> Deeper treatment: `2026-09-statistical-thinking.Rmd` runs both designs
> through full four-parameter-logistic fits and shows `alias()` returning the
> confounded term explicitly — the software telling you the design cannot
> answer the question.

### Takeaway

Good design decides which variation the statistics will be able to separate
later. After the experiment, no analysis can unmix a confounder.

---

# Module 5 — Comparing two conditions: paired or unpaired?

### Story

Here is the study as two clouds of dots — the way this comparison is usually
drawn.

```{r m5-clouds, fig.width=7, fig.height=4.2}
set.seed(105)  ## fixed so the deck and handout stay in sync
p5a <- dm %>%
  ggplot(aes(compound, viability, colour = compound)) +
  geom_point(size = 2.8, alpha = 0.85,
             position = position_jitter(width = 0.09, height = 0, seed = 7)) +
  stat_summary(fun = mean, geom = "crossbar", width = 0.45, colour = "grey25") +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "A  Two independent-looking groups",
       subtitle = "the clouds overlap substantially",
       x = NULL, y = "Viability (% of vehicle)")

p5b <- dm %>%
  ggplot(aes(compound, viability)) +
  geom_line(aes(group = donor_id), colour = "grey65", linewidth = 0.8) +
  geom_point(aes(colour = compound), size = 2.8, alpha = 0.9) +
  scale_colour_manual(values = cpd_cols, guide = "none") +
  coord_cartesian(ylim = c(0, 100)) +
  labs(title = "B  The same data, lots joined",
       subtitle = "every lot moves the same way",
       x = NULL, y = NULL)

p5a | p5b
```

> What information did the first graph hide?

Every lot was tested with **both** compounds. The vertical scatter in panel A
is mostly lots being different from one another, not compounds being similar.
Once the lines are drawn, the lot differences stop competing with the signal.

### Four tests, one dataset

```{r m5-tests}
wide <- dm %>%
  select(donor_id, compound, viability) %>%
  pivot_wider(names_from = compound, values_from = viability) %>%
  mutate(difference = `CPD-B` - `CPD-A`)

tests <- list(
  "Student t, unpaired (equal variance assumed)" =
    t.test(viability ~ compound, data = dm, var.equal = TRUE),
  "Welch t, unpaired (no equal-variance assumption)" =
    t.test(viability ~ compound, data = dm),
  "Paired t" =
    t.test(wide$`CPD-B`, wide$`CPD-A`, paired = TRUE),
  "Wilcoxon signed rank (paired, non-parametric)" =
    wilcox.test(wide$`CPD-B`, wide$`CPD-A`, paired = TRUE, conf.int = TRUE))

test_table <- imap_dfr(tests, function(tt, nm) {
  est <- if (grepl("^Paired|^Wilcoxon", nm)) unname(tt$estimate)
         else unname(-diff(rev(tt$estimate)))
  tibble(test = nm,
         estimate_pp = est,
         ci_low  = tt$conf.int[1] * if (grepl("unpaired", nm)) -1 else 1,
         ci_high = tt$conf.int[2] * if (grepl("unpaired", nm)) -1 else 1,
         p_value = tt$p.value)
}) %>%
  mutate(ci = pmin(ci_low, ci_high), ci_high = pmax(ci_low, ci_high),
         ci_low = ci,
         ci_width = ci_high - ci_low,
         beats_threshold = if_else(ci_low > MCID, "yes", "inconclusive")) %>%
  select(test, estimate_pp, ci_low, ci_high, ci_width, p_value,
         beats_threshold) %>%
  mutate(across(c(estimate_pp, ci_low, ci_high, ci_width), ~ round(.x, 1)),
         p_value = signif(p_value, 3))

export_data(test_table, "m05_four_tests",
            "the same comparison under four tests")
test_table
```

Read the last two columns, not the p-value column. Every test agrees there is
a difference. They do **not** agree on whether it clears the
`r round(MCID, 1)` pp threshold the team committed to in Module 1: the
unpaired analyses cannot tell, the paired analysis can. Ignoring the pairing
did not just cost significance — it cost the decision.

### The decision rule

Are observations linked by the experimental design?

* same lot under both compounds → **paired**
* same animal before and after → **paired**
* separately sampled groups → **unpaired**

Pairing is a property of the design, fixed before the data exist. It is never
something you discover by trying both and keeping the better result.

If the groups really are independent, do not assume equal variances without a
reason. **Welch's t-test** does not require equal variances, and when the
variances happen to be equal it costs almost nothing — which is exactly what
the first two rows of the table show here. The lot-level SDs of the two groups
are
`r paste(round(tapply(dm$viability, dm$compound, sd), 1), collapse = " and ")`
pp, close enough that Student and Welch agree to the first decimal. That is the
argument *for* Welch as the default: you get the protection when you need it
and pay nothing when you do not, and you avoid having to test for equal
variances first — a pre-test that itself has poor power at these sample sizes.

The Wilcoxon signed-rank test is the paired non-parametric option. On eight
pairs it gives a similar estimate with a larger p-value, because ranking eight
numbers throws away most of the information they contain. With *n* = 8 the
smallest p-value it can ever return is 1/128 ≈ 0.008 — so it cannot produce
strong evidence here no matter what the data look like. Non-parametric is not
a free lunch; it buys robustness to distributional assumptions with power.

### Visualisation hierarchy

1. the raw observations
2. the structure — pairing, blocks, batches
3. the estimated effect
4. its uncertainty

Bars, SEM whiskers and asterisks come after all four, if at all.

### Wide and long, the same 16 numbers

```{r m5-wide-long}
export_data(dm %>% select(donor_id, donor_sex, donor_age, weekday, compound,
                          viability, sd_wells),
            "m05_lot_means_long",
            "lot-level means, long format - one row per lot x compound (for R/Python)")
export_data(wide, "m05_lot_means_wide",
            "lot-level means, wide format - one row per lot (for Excel/Prism)")

wide %>% mutate(across(where(is.numeric), ~ round(.x, 1)))
```

The **long** format has one row per measurement and a column naming each
factor; it is what `ggplot2`, `dplyr` and most statistical models expect. The
**wide** format spreads one factor across columns; it is what spreadsheets and
Prism are built for, and it is the right shape for a paired test because the
pairing is carried by the row. Both files are exported. Converting between
them is `pivot_wider()` / `pivot_longer()`, and the conversion is lossless
only if the long table carries every identifier — which is a good reason to
keep lot, day and plate as real columns rather than as sheet names.

### Your turn (5 min)

Find an experiment in your own work where the samples are paired. Was it
analysed that way?

### Takeaway

The statistical test follows the experimental design. Draw the structure
before you test it, and the right test is usually obvious.

---

# Module 6 — P-values, effect sizes and confidence intervals

### Story

Assume, for a moment, that the two compounds are **identical**. Then run the
whole study — 8 lots, 3 wells, both compounds — `r params$n_sim` times in that
imaginary world, and record the difference each time.

```{r m6-null, fig.width=7, fig.height=4.2}
set.seed(106)  ## fixed so the deck and handout stay in sync
null_world <- replicate_studies(params$n_sim, n_donors = 8, n_wells = 3,
                                fold = 1)
observed   <- tests[["Paired t"]]$estimate

export_data(null_world %>% select(sim, estimate, p_value),
            "m06_null_simulation",
            sprintf("%d simulated studies in a world where the compounds are identical",
                    params$n_sim))

p6a <- ggplot(null_world, aes(estimate)) +
  geom_histogram(bins = 45, fill = "grey78", colour = "white") +
  geom_histogram(data = filter(null_world, abs(estimate) >= abs(observed)),
                 bins = 45, fill = "#c1443b", colour = "white") +
  geom_vline(xintercept = c(-1, 1) * observed, linetype = "dashed",
             linewidth = 1) +
  labs(title = "A  What 'no difference' looks like",
       subtitle = sprintf("dashed lines: the difference we actually observed (%.1f pp)",
                          observed),
       x = "Difference between compounds (pp)", y = "Simulated studies")

p6b <- ggplot(null_world, aes(p_value)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "grey78", colour = "white") +
  geom_histogram(data = filter(null_world, p_value < 0.05),
                 breaks = seq(0, 1, 0.05), fill = "#c1443b", colour = "white") +
  labs(title = "B  P-values when nothing is going on",
       subtitle = sprintf("flat, by construction; %.1f%% fall below 0.05",
                          100 * mean(null_world$p_value < 0.05)),
       x = "p-value", y = "Simulated studies")

p6a | p6b
```

Panel A *is* the p-value: the red area is the fraction of studies in a
no-difference world that would produce a difference at least as extreme as
ours. Here that fraction is
`r sprintf("%.4f", mean(abs(null_world$estimate) >= abs(observed)))`,
and the paired t-test reports
`r format.pval(tests[['Paired t']]$p.value, digits = 3)` — the same quantity, computed from
a formula instead of from `r params$n_sim` simulated experiments.

Panel B is worth a minute on its own. When there is genuinely nothing to find,
p-values are **uniform**: every value is as likely as every other, and 5% land
under 0.05. That is what a 5% false-positive rate means, and it is why a
single p-value below 0.05 is weak evidence on its own.

### What a p-value is not

A p-value is **not**:

* the probability that the compounds are identical
* the probability that the result happened by chance
* the probability that the experiment will replicate
* anything at all about the **size** of the effect

### Confidence intervals, and what "95%" refers to

```{r m6-coverage, fig.width=7, fig.height=5.4}
set.seed(107)  ## fixed so the deck and handout stay in sync
coverage_sims <- replicate_studies(100, n_donors = 8, n_wells = 3, fold = 5)
true_effect   <- mean(coverage_sims$estimate)

coverage_sims %>%
  mutate(covers = lwr <= true_effect & upr >= true_effect,
         sim = rank(estimate, ties.method = "first")) %>%
  ggplot(aes(sim, estimate, colour = covers)) +
  geom_hline(yintercept = true_effect, linetype = "dashed", linewidth = 1) +
  geom_linerange(aes(ymin = lwr, ymax = upr), linewidth = 0.8) +
  geom_point(size = 1.1) +
  coord_flip() +
  scale_colour_manual(values = c(`TRUE` = "grey55", `FALSE` = "#c1443b"),
                      labels = c(`TRUE` = "covers the truth",
                                 `FALSE` = "misses")) +
  labs(title = "One hundred repeats of the same study, one interval each",
       subtitle = "the truth is fixed; the intervals move",
       x = "Simulated study (sorted)", y = "Effect estimate (pp)") +
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        legend.position = "bottom")
```

"95% confidence" is a statement about the **procedure**, not about any one
interval: run the study forever and 95% of the intervals you build this way
will contain the truth. You never learn which kind you got. What the interval
does give you, immediately and usefully, is a range of effect sizes the data
are compatible with.

### Which result is more informative?

> **A.** p = 0.03
>
> **B.** estimated difference = 38 pp, plausible range 26 to 50 pp

Only B can be compared with a threshold. Only B supports a decision.

### Three results, three different conclusions

```{r m6-three, fig.width=7, fig.height=4.4}
three <- tribble(
  ~result,                     ~estimate, ~lwr, ~upr,
  "A  large, very uncertain",     45,      -5,   95,
  "B  tiny, very precise",         4,       1,    7,
  "C  moderate, moderate CI",     38,      26,   50) %>%
  mutate(verdict = case_when(
    lwr > MCID  ~ "relevant difference",
    upr < MCID  ~ "no relevant difference",
    TRUE        ~ "inconclusive"),
    verdict = factor(verdict, levels = c("relevant difference",
                                         "no relevant difference",
                                         "inconclusive")))

export_data(three, "m06_three_results",
            "three hypothetical results and what each licenses you to conclude")

ggplot(three, aes(estimate, fct_rev(result), colour = verdict)) +
  annotate("rect", xmin = -Inf, xmax = MCID, ymin = -Inf, ymax = Inf,
           fill = "grey92") +
  geom_vline(xintercept = MCID, linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = 0, colour = "grey60") +
  geom_linerange(aes(xmin = lwr, xmax = upr), linewidth = 1.2) +
  geom_point(size = 3) +
  annotate("text", x = MCID + 2, y = 0.56, hjust = 0, size = GEOM_PT,
           label = sprintf("threshold %.1f pp", MCID)) +
  scale_x_continuous(expand = expansion(mult = c(0.03, 0.07))) +
  guides(colour = guide_legend(nrow = 2)) +
  scale_colour_manual(values = c("relevant difference" = "#3b7a57",
                                 "no relevant difference" = "#4c8fbd",
                                 "inconclusive" = "#d9a441")) +
  labs(title = "Read the interval against the threshold, not against zero",
       subtitle = "grey: differences too small to act on",
       x = "Difference between compounds (pp)", y = NULL) +
  theme(legend.position = "bottom")
```

B is the interesting one: a precise, unambiguous, *statistically significant*
effect that is far too small to matter. Statistical significance and biological
importance are different questions, and only the second one was ever the
reason for running the assay.

### Three conclusions, not two

Most people are trained to report two outcomes. There are three:

1. **Evidence for a relevant difference** — the interval lies above the
   threshold
2. **Evidence against a relevant difference** — the interval lies entirely
   below it
3. **Inconclusive** — the interval spans it; the experiment cannot separate
   the two, and that is a result worth reporting

### Our study

```{r m6-verdict}
paired  <- tests[["Paired t"]]
verdict <- tibble(
  estimate_pp = paired$estimate,
  ci_low      = paired$conf.int[1],
  ci_high     = paired$conf.int[2],
  threshold   = MCID,
  p_value     = paired$p.value,
  conclusion  = case_when(
    paired$conf.int[1] > MCID ~ "relevant difference - act",
    paired$conf.int[2] < MCID ~ "no relevant difference - do not act",
    TRUE                      ~ "inconclusive - more lots needed")) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

export_data(verdict, "m06_study_verdict",
            "the pre-registered decision applied to our data")
verdict
```

### Your turn (5 min)

Take the three results in the figure above. Which of them would change what
you do next, and why? Then think of the last p-value you reported in a
presentation: could you still state the effect size and its interval? If not,
what did your audience actually take away?

### Takeaway

The p-value answers one narrow question about a world we do not believe in.
The effect size and its interval answer the question we asked. Report both,
decide on the second.

---

# Module 7 — Surprising observations

### Story

Two values in the dataset look wrong. They are not the same kind of thing.

```{r m7-candidates, fig.width=7, fig.height=5}
lot_flags <- viab %>%
  group_by(donor_id, compound) %>%
  mutate(well_z = (viability_pct - median(viability_pct)) /
                  (mad(viability_pct) + 1e-9)) %>%
  ungroup()

export_data(lot_flags %>% select(donor_id, compound, weekday, replicate,
                                 viability_pct, well_z),
            "m07_outlier_candidates",
            "all wells with a within-condition robust z score")

lot_flags %>%
  ggplot(aes(fct_reorder(donor_id, viability_pct), viability_pct,
             colour = compound)) +
  geom_point(size = 2.4, alpha = 0.85,
             position = position_dodge(width = 0.5)) +
  ## the bubble well is the CPD-B point of HH2, dodged to the right of centre
  annotate("curve", x = 5.75, xend = 4.3, y = 13, yend = 20.3, curvature = -0.3,
           arrow = arrow(length = unit(0.18, "cm")), colour = "grey35") +
  annotate("text", x = 8.45, y = 9, hjust = 1, size = GEOM_PT, colour = "grey25",
           label = "candidate 1: one well,\nfar from its two siblings") +
  annotate("curve", x = 1.6, xend = 0.92, y = 66, yend = 45, curvature = 0.3,
           arrow = arrow(length = unit(0.18, "cm")), colour = "grey35") +
  annotate("text", x = 1.7, y = 72, hjust = 0, size = GEOM_PT, colour = "grey25",
           label = "candidate 2: a whole lot,\nlow for BOTH compounds") +
  scale_colour_manual(values = cpd_cols) +
  labs(title = "All 48 wells",
       subtitle = "one of these is an artefact and one is biology",
       x = "Hepatocyte lot", y = "Viability (% of vehicle)") +
  coord_cartesian(ylim = c(0, 104)) +
  theme(legend.position = "bottom")
```

### Candidate 1 — a single well

```{r m7-well}
lot_flags %>%
  filter(donor_id == sim$params$bubble_donor,
         compound == sim$params$bubble_compound) %>%
  select(donor_id, compound, replicate, raw_rlu, viability_pct, well_z)
```

The laboratory notebook for that plate says:

> *Air bubble observed in well B; pipette felt strange.*

An independent, documented, technical cause, recorded **before** anyone looked
at the result. The other two wells of the same condition agree with each
other. This is a measurement failure, and excluding it is a QC decision, not
a statistical one.

### Candidate 2 — a whole lot

```{r m7-grubbs}
grubbs <- function(x) {
  x <- x[!is.na(x)]; n <- length(x)
  G     <- max(abs(x - mean(x))) / sd(x)
  tcrit <- qt(0.05 / (2 * n), df = n - 2)
  Gcrit <- (n - 1) / sqrt(n) * sqrt(tcrit^2 / (n - 2 + tcrit^2))
  tibble(n = n, G = round(G, 3), G_crit = round(Gcrit, 3), flagged = G > Gcrit)
}

dm %>%
  group_by(compound) %>%
  group_modify(~ grubbs(.x$viability)) %>%
  ungroup()
```

Lot `r sim$params$sensitive_donor` is more sensitive to **both** compounds —
its IC~50~ is genuinely about
`r sim$params$sensitive_fold`-fold lower. A routine outlier test will flag it
anyway, because it is answering the wrong question: Grubbs asks whether a value
is improbable under an assumed distribution, never whether it is *wrong*.
Nothing in the QC record marks that plate. A lot that is genuinely more
sensitive is precisely the observation a safety margin should be built on.

### A workflow, not a test

1. **Verify** — transcription, instrument, pipetting, sample identity
2. **Investigate** — is there an experimental explanation in the record?
3. **Apply pre-defined rules** — was an exclusion criterion fixed before the
   result was seen?
4. **Do not delete because it is inconvenient** — biology is often surprising
5. **Run a sensitivity analysis** — does the conclusion change either way?
6. **Document** — report what was excluded and why

### Step 5, done

```{r m7-sensitivity}
no_bubble <- viab %>% filter(!bubble) %>% donor_means()
no_lot    <- dm %>% filter(donor_id != sim$params$sensitive_donor)

run_paired <- function(d, label) {
  w  <- d %>% select(donor_id, compound, viability) %>%
    pivot_wider(names_from = compound, values_from = viability)
  tt <- t.test(w$`CPD-B`, w$`CPD-A`, paired = TRUE)
  tibble(scenario = label, n_lots = nrow(w),
         estimate_pp = tt$estimate, ci_low = tt$conf.int[1],
         ci_high = tt$conf.int[2], p_value = tt$p.value)
}

sensitivity <- bind_rows(
  run_paired(dm,        "keep everything"),
  run_paired(no_bubble, "exclude the air-bubble well (documented QC failure)"),
  run_paired(no_lot,    "exclude the sensitive lot (no technical justification)"),
  run_paired(no_bubble %>% filter(donor_id != sim$params$sensitive_donor),
             "exclude both")) %>%
  mutate(clears_threshold = if_else(ci_low > MCID, "yes", "inconclusive"),
         across(c(estimate_pp, ci_low, ci_high), ~ round(.x, 1)),
         p_value = signif(p_value, 3))

export_data(sensitivity, "m07_sensitivity_analysis",
            "the conclusion under four exclusion scenarios")
sensitivity
```

Dropping the bubble well changes the estimate barely at all — because it was
one of three wells that were then averaged, and averaging is already doing
most of the work. Dropping the sensitive lot is a different matter: the
pairing had absorbed that lot completely (it is low for both compounds, so its
*difference* is ordinary), so removing it discards a lot that was costing
nothing and carrying real information about how variable human hepatocytes are.

> The companion document extends this to a whole plate that fails
> pre-specified QC, with Z' factor, vehicle CV and signal window as the
> criteria, and to the case where you should model the extra variance rather
> than delete it.

### Your turn (5 min)

Think of a value you or a colleague once excluded from an analysis. Which of
the six steps above justified it? Was the criterion written down before the
result was seen — and would the conclusion have survived if the value had
stayed in?

### Takeaway

Ask which question you are answering: is the value *implausible given the
controls*, or merely *inconvenient given the hypothesis*? The first is a QC
exclusion. The second is data manipulation. Running it both ways is cheap and
almost always the right thing to report.

---

# Module 8 — Power, sample size and the decision

### Story

Back to the original question, with the whole dataset visible.

> Would you change the project's plan on the basis of these data? What
> experiment, if any, would you run next?

### Power, without formulas

The same true `r thr$true_fold`-fold difference, the same assay, the same
noise. Only the number of lots changes.

```{r m8-power, fig.width=7, fig.height=4.5}
set.seed(108)  ## fixed so the deck and handout stay in sync
power_grid <- map_dfr(c(3, 6, 12, 24), function(n)
  replicate_studies(params$n_sim, n_donors = n, n_wells = 3,
                    fold = thr$true_fold))

power_summary <- power_grid %>%
  group_by(n_donors) %>%
  summarise(
    power_significant = mean(significant),
    power_decisive    = mean(calls_relevant),
    median_estimate   = median(estimate),
    median_ci_width   = median(upr - lwr),
    .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))

export_data(power_summary, "m08_power_by_n",
            "power to reach significance vs power to clear the relevance threshold")
power_summary
```

```{r m8-power-fig, fig.width=7, fig.height=8.4}
p8a <- power_summary %>%
  pivot_longer(c(power_significant, power_decisive),
               names_to = "criterion", values_to = "power") %>%
  mutate(criterion = recode(criterion,
           power_significant = "p < 0.05\n(\"is there any difference?\")",
           power_decisive    = "CI above the threshold\n(\"is it big enough to act on?\")")) %>%
  ggplot(aes(n_donors, power, colour = criterion)) +
  geom_hline(yintercept = 0.8, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 1.1) + geom_point(size = 2.6) +
  scale_x_continuous(breaks = c(3, 6, 12, 24)) +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1)) +
  scale_colour_manual(values = c("#4c8fbd", "#3b7a57")) +
  labs(title = "A  Two kinds of power",
       subtitle = "getting a small p-value is much easier than answering the question",
       x = "Independent hepatocyte lots", y = "Proportion of studies") +
  theme(legend.position = "bottom")

p8b <- power_grid %>%
  mutate(n_lab = factor(paste0("n = ", n_donors),
                        levels = paste0("n = ", c(3, 6, 12, 24)))) %>%
  ggplot(aes(estimate, n_lab, fill = n_lab)) +
  geom_violin(colour = NA, alpha = 0.75) +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white") +
  geom_vline(xintercept = median(power_grid$estimate[power_grid$n_donors == 24]),
             linetype = "dashed", linewidth = 1) +
  scale_fill_brewer(palette = "Blues", guide = "none") +
  labs(title = "B  The estimate itself, not just its p-value",
       subtitle = "at n = 3 the study can return almost any answer",
       x = "Estimated difference (pp)", y = NULL)

p8a / p8b
```

Two things to take from panel A. First, power is not a property of an assay —
it is a property of an assay **plus a hypothesised effect size**. "How many
samples do I need?" has no answer until "what difference would matter?" has
one. Second, the two curves are far apart: with
`r power_summary$n_donors[1]` lots you would reach significance in
`r scales::percent(power_summary$power_significant[1])` of studies but could only
*act* on `r scales::percent(power_summary$power_decisive[1])` of them.

Panel B is the reason small studies are dangerous rather than merely weak. At
*n* = 3 the same experiment, run on the same compounds, can return almost any
answer.

There is a second, less obvious cost. A study that is underpowered can only
reach significance by landing on an unusually large estimate — so among the
studies that *do* get published, the effect is systematically overstated. The
5-fold difference above is too easy to detect for this to bite, so the
simulation below uses a genuinely marginal effect instead.

```{r m8-exaggeration, fig.width=7, fig.height=4.5}
set.seed(109)  ## fixed so the deck and handout stay in sync
marginal <- map_dfr(c(3, 4, 6, 12, 24), function(n)
  replicate_studies(params$n_sim, n_donors = n, n_wells = 3, fold = 1.3))

exaggeration <- marginal %>%
  group_by(n_donors) %>%
  summarise(power            = mean(significant),
            all_studies      = median(estimate),
            significant_only = median(estimate[significant]),
            .groups = "drop") %>%
  mutate(exaggeration = significant_only / all_studies,
         across(where(is.numeric), ~ round(.x, 2)))

export_data(exaggeration, "m08_exaggeration",
            "with a marginal true effect: median estimate among all studies vs among significant ones")
exaggeration
```

```{r m8-exaggeration-fig, fig.width=7, fig.height=4.2}
ggplot(exaggeration, aes(power, exaggeration)) +
  geom_hline(yintercept = 1, linetype = "dashed", colour = "grey45") +
  geom_line(linewidth = 1.2, colour = "#c1443b") +
  geom_point(size = 2.8, colour = "#c1443b") +
  geom_text(aes(label = paste0("n = ", n_donors)), vjust = -1.2, size = GEOM_PT) +
  scale_x_continuous(labels = scales::percent) +
  expand_limits(y = c(0.95, 1.5)) +
  labs(title = "The less power a study has, the more it exaggerates",
       subtitle = "ratio of the median published effect to the median true-study effect, when only significant studies are reported",
       x = "Power", y = "Exaggeration of the reported effect")
```

At `r scales::percent(exaggeration$power[1])` power the surviving studies
overstate the effect by about
`r scales::percent(exaggeration$exaggeration[1] - 1)`; by the time power
reaches `r scales::percent(exaggeration$power[nrow(exaggeration)])` the
distortion is gone. An underpowered literature is not merely a noisy one — it
is a biased one, and the bias always points the same way.

### Where to spend twenty-four wells

```{r m8-allocation, fig.width=7, fig.height=4.2}
set.seed(110)  ## fixed so the deck and handout stay in sync
allocation <- bind_rows(
  replicate_studies(params$n_sim, n_donors = 3, n_wells = 8,
                    fold = thr$true_fold) %>%
    mutate(allocation = "3 lots × 8 wells"),
  replicate_studies(params$n_sim, n_donors = 8, n_wells = 3,
                    fold = thr$true_fold) %>%
    mutate(allocation = "8 lots × 3 wells"))

allocation %>%
  group_by(allocation) %>%
  summarise(power_significant = mean(significant),
            power_decisive    = mean(calls_relevant),
            median_ci_width   = median(upr - lwr), .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  export_data("m08_allocation",
              "24 wells spent as 3 lots x 8 wells or 8 lots x 3 wells")

allocation %>%
  ggplot(aes(estimate, fill = allocation)) +
  geom_density(alpha = 0.55, colour = NA) +
  geom_vline(xintercept = MCID, linetype = "dashed", linewidth = 1) +
  annotate("text", x = MCID + 1, y = 0, hjust = 0, vjust = -0.6, size = GEOM_PT,
           label = "relevance threshold") +
  scale_fill_manual(values = c("#c1443b", "#3b7a57")) +
  labs(title = "The same 24 wells, allocated two ways",
       subtitle = "technical replicates cannot substitute for independent lots",
       x = "Estimated difference (pp)", y = "Density") +
  theme(legend.position = "bottom")
```

Both use 24 wells per compound. Three lots with eight wells each measures three
lots very precisely and then has to generalise from three; eight lots with
three wells each measures each lot a little less precisely and generalises from
eight. Since the uncertainty that matters is lot-to-lot, the second wins — and
it keeps winning no matter how many wells you add to the first.

### Your turn (5 min)

Would you act on these data? What would you run next, and how many lots would
it need?

### The final checklist

1. What exactly is the question, and what difference would matter?
2. What population do I want to generalise to?
3. What is my independent experimental unit?
4. Where can bias enter?
5. Which variation do I block, randomise, or standardise?
6. Plot the observations so the experimental structure is visible.
7. Estimate the effect **and** its uncertainty.
8. Use a test that matches the design.
9. Investigate surprising observations instead of deleting them.
10. Decide using effect size, uncertainty and context — not the p-value alone.

### Takeaway

Power is not a number you compute at the end to justify what you did. It is
the question "what would this experiment be able to tell me?", asked while
there is still time to change the answer.

---

# Module 9 — What we have learned

One dataset, ten questions, and one answer that changed as we asked better
questions about the design:

```{r m9-journey}
journey <- bind_rows(
  tibble(step = "1. count wells, ignore lots (Module 3)",
         estimate_pp = pseudo$estimate_pp[1],
         ci_low = pseudo$ci_low[1], ci_high = pseudo$ci_high[1]),
  tibble(step = "2. confounded schedule (Module 4)",
         estimate_pp = design_summary$mean_difference_pp[
           design_summary$design == "Confounded"],
         ci_low = NA_real_, ci_high = NA_real_),
  tibble(step = "3. lots, but pairing ignored (Module 5)",
         estimate_pp = test_table$estimate_pp[2],
         ci_low = test_table$ci_low[2], ci_high = test_table$ci_high[2]),
  tibble(step = "4. lots, paired, blocked (Modules 5-7)",
         estimate_pp = test_table$estimate_pp[3],
         ci_low = test_table$ci_low[3], ci_high = test_table$ci_high[3])) %>%
  mutate(threshold_pp = round(MCID, 1),
         conclusion = case_when(
           is.na(ci_low)     ~ "no interval - the design cannot separate day from compound",
           ci_low > MCID     ~ "relevant difference",
           ci_high < MCID    ~ "no relevant difference",
           TRUE              ~ "inconclusive"),
         earned = c("no - the interval is too narrow to be believed",
                    "no - the estimate contains the run-day effect",
                    "yes",
                    "yes"),
         across(where(is.numeric), ~ round(.x, 1)))

export_data(journey, "m09_summary_of_analyses",
            "the same 48 numbers analysed four ways")
journey
```

Same 48 numbers every time. Only step 4 is both right and *entitled* to be
right.

Step 1 is the uncomfortable one: counting wells instead of lots reaches the
same conclusion as the correct analysis, with a far smaller p-value. It is
right by accident. Had the lots been a little more variable, or the pairing a
little weaker, the same shortcut would have produced the same unwarranted
confidence behind a wrong answer — and nothing in the output would have looked
different. This is why the questions in this workshop are about the design
rather than about the arithmetic: the arithmetic always runs, and it never
tells you which of these four rows you are in.

### The three things worth remembering

* Good design **minimises unnecessary and avoidable variation** while
  maximising the chance of detecting the effect you care about. It decides
  what the statistics will be able to separate later.
* Design and analysis **reduce uncertainty; they never remove it.** An honest
  interval is a better deliverable than a confident sentence.
* The decision is **scientific**. Effect size, threshold, context and prior
  knowledge live outside statistics; the analysis only tells you how much of
  the data's uncertainty you have to carry into that decision.

### The one-line version

> To answer *"which statistical test should I use?"* you first have to answer
> *"how were the data generated?"*

---

# Module 10 — Where to go from here

### Getting help from an LLM

This is where the workshop pays off in practice. Consider the usual prompt:

> *"Compare columns A and B statistically."*

An LLM will answer it. It will pick a test, report a p-value, and sound
confident. It cannot possibly know whether the columns are paired, whether the
rows are wells or lots, whether the two groups were run on different days, or
what difference you would act on. So it will guess, and the guess is usually
an unpaired t-test on every row — which, on this dataset, is exactly the
analysis that gets the wrong answer twice over.

Now the same request with the design in it:

> *"These measurements come from eight independent human hepatocyte lots. Each
> lot was tested with both compounds, with three technical replicate wells per
> lot and compound. The independent experimental unit is the lot, not the
> well. The plates were run over two days; day is balanced across compounds.
> We decided in advance that a difference smaller than 24 percentage points of
> viability would not change our decision. Help me visualise the data so the
> pairing is visible, estimate the compound difference with its uncertainty,
> and compare that interval with our threshold. Before you recommend a test,
> tell me what you think the independent unit is and whether the comparison is
> paired."*

Everything that makes the second prompt work is a design fact, and every one
of those facts was decided before any data existed. The final sentence is the
most useful habit: **ask the model to state its assumptions about the unit and
the pairing before it computes anything**, so that you can catch a wrong
assumption while it is still cheap.

The same applies to a colleague, a statistician, or your own notebook six
months from now.

### Data to take away

Every module exported the data it used. They are in
`r params$outdir`, as plain CSV — open them in Excel, Prism, R, Python, or
paste one into an LLM along with the design description above.

```{r m9-manifest}
.manifest %>% arrange(file)
```

```{r m9-full-exports}
## the complete dataset in both shapes, plus the hidden truth
export_data(viab, "full_dataset_long",
            "all 48 wells with plate references (long format)")

export_data(
  viab %>%
    select(donor_id, donor_sex, donor_age, weekday, compound, replicate,
           viability_pct) %>%
    pivot_wider(names_from = c(compound, replicate),
                values_from = viability_pct,
                names_glue = "{compound}_well{replicate}"),
  "full_dataset_wide",
  "all 48 wells, one row per lot (wide format, for Excel/Prism)")

export_data(
  sim$truth %>%
    select(donor_id, compound, ic50_true, viab_true, sensitive) %>%
    mutate(across(where(is.numeric), ~ round(.x, 2))),
  "answer_key_true_values",
  "the simulated ground truth - the answer key, for after the workshop")
```

### Reading

* Krzywinski & Altman, *Points of Significance*, Nature Methods (2013–2015) —
  short, visual, and written for exactly this audience
* Lazic, *Experimental Design for Laboratory Biologists* (2016) — the best
  single book on pseudoreplication and blocking in bench science
* Amrhein, Greenland & McShane, *Retire statistical significance*, Nature
  (2019) — the argument behind Module 6
* The ARRIVE and *Nature* reporting checklists — useful even when you are not
  publishing

### And otherwise

Come and talk to me. Most statistical problems that reach a statistician are
design problems that arrived too late, and the conversation is much shorter
and much happier when it happens **before** the experiment.

---

# Appendix — the companion document

`2026-09-statistical-thinking.Rmd` takes the same assay to full analyst depth
and is the place to look when a module raised a question this notebook
deliberately simplified:

| Topic | Where |
|---|---|
| Full 96-well plate layout, edge effects | *Design* |
| Eight concentrations, four-parameter logistic fits | *Curve fitting* |
| Absolute vs relative IC~50~, and why the difference matters | *Curve fitting* |
| Pre-specified plate QC: Z', vehicle CV, signal window | *Plate quality control* |
| `alias()` and formally undiagnosable designs | *Consequences of the two designs* |
| A whole plate failing technically | *Perturbation 1* |
| Grubbs vs a genuinely sensitive lot | *Perturbation 2* |
| Weighted fitting, biological vs technical variance | *Perturbation 3* |

# Session information

```{r session}
sessionInfo()
```
