---
title: "CellTiter-Glo in primary human hepatocytes: two compounds, two run days"
subtitle: "Run-day confounding, a failed plate, a sensitive donor, and unequal technical variance"
author: "Synthetic data — for method illustration only"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    fig_width: 7
    fig_height: 4.5
params:
  seed: 20260919
  ic50_A: 10       # uM, ground truth for CPD-A (the more cytotoxic compound)
  ic50_B: 50       # uM, ground truth for CPD-B
  hill_A: 1.7
  hill_B: 1.5
  gsd_A: 1.35      # donor-to-donor geometric SD of the IC50
  gsd_B: 1.45
  friday_vehicle_gain: 0.20  # vehicle wells read 20% higher on Friday
  edge_loss: 0.10  # fractional signal loss in perimeter wells; 0 switches it off
  ## --- perturbation 1: one plate fails technically -------------------------
  failed_donor: "HH4"
  failed_compound: "CPD-B"
  failed_gain: 0.15     # whole-plate signal collapse (reagent under-dispense)
  failed_extra_cv: 0.30 # additional well-to-well CV on that plate
  ## --- perturbation 2: one genuinely sensitive donor -----------------------
  sensitive_donor: "HH7"
  sensitive_factor: 0.70  # IC50 30% lower for BOTH compounds - real, not an error
  ## --- perturbation 3: unequal technical variance --------------------------
  floor_cv_gain: 0.12   # extra CV as the signal approaches the assay floor
  outdir: "output"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.align = "center", dev = "png", dpi = 200,
                      ## every figure is drawn at its final size - 7 inches or
                      ## less - so that 12 pt type really is 12 pt on the page
                      fig.width = 7, fig.height = 4.5)
```

# Purpose

Two compounds are compared for cytotoxicity in primary human hepatocytes using
the CellTiter-Glo ATP readout. Ground truth is fixed at IC~50~ =
`r params$ic50_A` µM for CPD-A and `r params$ic50_B` µM for CPD-B — a true
`r round(params$ic50_B / params$ic50_A, 1)`-fold separation.

The plates cannot all be run on one day, so the assay is split over **Friday
and Monday**. The simulated truth is that the *vehicle wells* read
`r scales::percent(params$friday_vehicle_gain)` higher on Friday, while the
compound-treated wells are unchanged. Because viability is expressed relative
to the plate's own vehicle, a day effect that lives only in the reference
propagates straight into every normalised value on that plate.

Two designs are simulated on the same plate layout, differing **only** in how
run day is allocated:

* **Confounded.** All CPD-A plates run on Friday, all CPD-B plates
  on Monday. This is the realistic version of the mistake: the compounds are on
  separate plates, so nothing stops them being scheduled on separate days.
* **Blocked.** Run day is allocated by donor: both compounds of a
  given donor are run on the same day, with four donors on Friday and four on
  Monday. Compound is then orthogonal to day.

# Packages

```{r libraries}
library(tidyverse)
library(broom)
library(patchwork)

library(ribiosUtils)  # wellIndex2position, equateWellLabelWidth, ofactor, bound
library(ribiosPlot)   # royalbluered, p2asterisk, openFileDevice, closeFileDevice

set.seed(params$seed)
dir.create(params$outdir, showWarnings = FALSE, recursive = TRUE)

## One theme for the whole document. Nothing below 12 pt, and every figure is
## drawn at 7 inches wide or less, so the text keeps a generous share of the
## canvas rather than shrinking when the figure is scaled onto a page.
BASE_PT <- 12                  # smallest type anywhere in a figure
GEOM_PT <- BASE_PT / .pt       # the same size for geom_text(), in mm

theme_ctg <- theme_bw(base_size = BASE_PT) +
  theme(plot.title    = element_text(face = "bold", size = BASE_PT + 1.5),
        plot.subtitle = element_text(size = BASE_PT, colour = "grey30"),
        axis.title    = element_text(size = BASE_PT),
        axis.text     = element_text(size = BASE_PT),
        strip.text    = element_text(size = BASE_PT, face = "bold"),
        legend.text   = element_text(size = BASE_PT),
        legend.title  = element_blank(),
        plot.title.position = "plot",
        legend.key.size = unit(0.9, "lines"),
        plot.margin   = margin(4, 6, 4, 4),
        panel.grid.minor = element_blank())

theme_ctg_flat <- theme_ctg + theme(panel.grid = element_blank())
theme_set(theme_ctg)
```

# Design

## Donors

```{r donors}
donors <- tibble(
  donor_id    = c("HH1", "HH2", "HH3", "HH4",
                  "HH5", "HH6", "HH7", "HH8"),
  donor_sex   = c("F", "M", "F", "M", "M", "F", "F", "M"),
  donor_age   = c(34L, 52L, 61L, 45L, 29L, 58L, 41L, 66L),
  vehicle_rlu = c(2.10, 1.68, 2.44, 1.92, 2.28, 1.55, 2.02, 1.79) * 1e6,
  blank_rlu   = 1.2e4
) %>%
  mutate(donor_id = ofactor(donor_id))

donors
```

## Plate layout: one compound per plate

This is the change that makes the confounding possible. With both compounds on
one plate they are unavoidably run together; splitting them onto separate
plates is what lets a scheduler put them on different days.

| Columns | Content |
|---|---|
| 1 and 11 | vehicle, 0.1% DMSO |
| 2 | lysis control, 100 µM digitonin (defines the assay floor) |
| 3–10 | eight concentrations, 0.3 → 1000 µM |
| 12 | no-cell blank (reagent + medium), defines the background |

Rows A–H are eight replicate wells of the single compound on that plate.

```{r plate-design}
conc_series <- c(0.3, 1, 3, 10, 30, 100, 300, 1000)

plate_design <- wellIndex2position(1:96, format = "96") %>%
  as_tibble() %>%
  rename(well_index = WellIndex, row = Row, column = Column) %>%
  mutate(
    well      = equateWellLabelWidth(paste0(row, column)),
    is_edge   = row %in% c("A", "H") | column %in% c(1L, 12L),
    well_type = case_when(
      column %in% c(1L, 11L) ~ "vehicle",
      column == 2L           ~ "lysis control",
      column == 12L          ~ "no-cell blank",
      TRUE                   ~ "test article"),
    replicate  = if_else(well_type == "test article",
                         match(row, LETTERS), NA_integer_),
    conc_index = if_else(column >= 3L & column <= 10L, column - 2L, NA_integer_),
    conc_uM    = case_when(
      well_type == "test article"  ~ conc_series[conc_index],
      well_type == "lysis control" ~ 100,
      TRUE                         ~ 0),
    row = ofactor(row))

count(plate_design, well_type)
```

```{r plate-layout-view, fig.width=7, fig.height=3.6}
layout_cols <- c("test article" = "grey92", "vehicle" = "#cfe3f5",
                 "lysis control" = "#f7cfcb", "no-cell blank" = "#f3e2b0")

p_layout <- plate_design %>%
  mutate(label = case_when(
    well_type == "test article"  ~ format(conc_uM, trim = TRUE,
                                          drop0trailing = TRUE,
                                          scientific = FALSE),
    well_type == "lysis control" ~ "dig",
    well_type == "vehicle"       ~ "veh",
    TRUE                         ~ "blk")) %>%
  ggplot(aes(x = factor(column), y = fct_rev(row))) +
  geom_tile(aes(fill = well_type), colour = "white", linewidth = 1.1) +
  geom_text(aes(label = label), size = GEOM_PT * 0.72) +
  scale_fill_manual(values = layout_cols, name = NULL) +
  labs(title = "Plate layout — one compound per plate, eight replicate rows",
       subtitle = "numbers are concentrations in µM",
       x = "Column", y = "Row") +
  theme_ctg_flat + theme(legend.position = "bottom")

p_layout
```

## Allocating the two run days

Sixteen plates: eight donors × two compounds. Only the `weekday` column differs
between the designs.

```{r day-allocation}
plates <- expand_grid(donor_id = donors$donor_id,
                      compound = c("CPD-A", "CPD-B")) %>%
  left_join(donors, by = "donor_id") %>%
  mutate(donor_index = as.integer(donor_id))

DESIGNS <- c("Blocked", "Confounded")   # referred to by name throughout

schedule <- bind_rows(
  ## day allocated by donor: both compounds of a donor run together
  plates %>% mutate(design  = DESIGNS[1],
                    weekday = if_else(donor_index <= 4L, "Friday", "Monday")),
  ## day allocated by compound: perfectly aliased
  plates %>% mutate(design  = DESIGNS[2],
                    weekday = if_else(compound == "CPD-A", "Friday", "Monday"))
) %>%
  mutate(design   = factor(design, levels = DESIGNS),
         weekday  = factor(weekday, levels = c("Friday", "Monday")),
         plate_id = sprintf("%s_%s_%s", substr(as.character(design), 1, 1),
                            str_remove(as.character(donor_id), "HH"),
                            str_remove(compound, "CPD-")))

schedule %>% count(design, compound, weekday)
```

```{r design-view, fig.width=7, fig.height=4}
day_cols <- c("Friday" = "#e07b39", "Monday" = "#4c8fbd")

p_design <- schedule %>%
  ggplot(aes(x = compound, y = fct_rev(donor_id), fill = weekday)) +
  geom_tile(colour = "white", linewidth = 1.4) +
  geom_text(aes(label = str_sub(weekday, 1, 3)), size = GEOM_PT * 0.85,
            colour = "white", fontface = "bold") +
  facet_wrap(~ design) +
  scale_fill_manual(values = day_cols, name = NULL) +
  labs(title = "Run-day allocation — the only difference between the designs",
       subtitle = "left: day balanced across compounds · right: day perfectly aliased with compound",
       x = NULL, y = "Donor") +
  theme_ctg_flat + theme(legend.position = "bottom")

p_design
```

# Simulation

Each donor draws its own IC~50~ around the reference value, log-normally. The
same donor-level truth is used for both designs, so any difference between them
comes from the day allocation alone.

```{r truth}
truth <- expand_grid(donor_id = donors$donor_id,
                     compound = c("CPD-A", "CPD-B")) %>%
  mutate(
    ic50_ref  = if_else(compound == "CPD-A", params$ic50_A, params$ic50_B),
    gsd       = if_else(compound == "CPD-A", params$gsd_A,  params$gsd_B),
    hill      = if_else(compound == "CPD-A", params$hill_A, params$hill_B),
    ic50_true = ic50_ref * rlnorm(n(), meanlog = 0, sdlog = log(gsd)),
    ## perturbation 2: one donor is genuinely more sensitive to both compounds
    sensitive = as.character(donor_id) == params$sensitive_donor,
    ic50_true = if_else(sensitive, ic50_true * params$sensitive_factor, ic50_true))

truth %>%
  select(donor_id, compound, ic50_true) %>%
  pivot_wider(names_from = compound, values_from = ic50_true) %>%
  mutate(across(where(is.numeric), ~ round(.x, 1)))
```

The day effect enters in exactly one place: on a Friday plate the **vehicle**
wells are multiplied by `r 1 + params$friday_vehicle_gain`. Compound-treated
wells, the lysis control and the blank are untouched.

Three further sources of trouble are built in, all of them things that happen:

1. **A technically failed plate** — `r params$failed_compound` on donor
   `r params$failed_donor` receives too little reagent, so the whole plate reads
   at `r scales::percent(params$failed_gain)` of normal with a much larger
   well-to-well spread.
2. **A genuinely sensitive donor** — `r params$sensitive_donor` has IC~50~
   values `r scales::percent(1 - params$sensitive_factor)` lower for *both*
   compounds. This is biology, not an error.
3. **Unequal technical variance** — hepatocyte lots differ in plating
   uniformity, so the well-to-well CV differs by donor; on top of that, the CV
   grows as the signal approaches the assay floor.

```{r noise-model}
donor_noise <- tibble(
  donor_id = donors$donor_id,
  tech_cv  = c(0.045, 0.050, 0.115, 0.055, 0.060, 0.105, 0.050, 0.048))

donor_noise %>% mutate(tech_cv = percentage(tech_cv))
```

```{r simulate}
wells <- schedule %>%
  select(design, plate_id, donor_id, donor_sex, donor_age, compound, weekday,
         vehicle_rlu, blank_rlu) %>%
  expand_grid(plate_design) %>%
  left_join(truth %>% select(donor_id, compound, ic50_true, hill),
            by = c("donor_id", "compound")) %>%
  left_join(donor_noise, by = "donor_id") %>%
  mutate(
    viab_true = case_when(
      well_type == "test article" ~ 1 / (1 + (conc_uM / ic50_true)^hill),
      well_type == "vehicle"      ~ 1,
      TRUE                        ~ 0),
    ## the day effect lives only in the vehicle wells
    day_gain = if_else(well_type == "vehicle" & weekday == "Friday",
                       1 + params$friday_vehicle_gain, 1),
    edge_factor = if_else(is_edge, 1 - params$edge_loss, 1),
    ## perturbation 1: whole-plate technical failure
    failed_plate = as.character(donor_id) == params$failed_donor &
                   compound == params$failed_compound,
    plate_gain = if_else(failed_plate, params$failed_gain, 1),
    ## perturbation 3: CV depends on the lot and rises towards the assay floor
    well_cv = tech_cv + params$floor_cv_gain * (1 - viab_true) +
              if_else(failed_plate, params$failed_extra_cv, 0),
    raw_rlu = if_else(
      well_type == "no-cell blank",
      blank_rlu * plate_gain * rnorm(n(), 1, 0.10),
      (vehicle_rlu * day_gain * edge_factor * viab_true * rnorm(n(), 1, well_cv) +
         blank_rlu * rnorm(n(), 1, 0.08)) * plate_gain),
    raw_rlu = round(pmax(raw_rlu, 0)))
```

## Normalisation

Per plate: background from the no-cell blanks, 100% from the **interior**
vehicle wells. This is the correct procedure — and it is precisely what
transmits the day effect, because the reference itself has moved.

```{r normalise}
ctg <- wells %>%
  group_by(design, plate_id) %>%
  mutate(
    plate_blank_rlu        = median(raw_rlu[well_type == "no-cell blank"]),
    plate_vehicle_interior = median(raw_rlu[well_type == "vehicle" & !is_edge]),
    viability_pct = 100 * (raw_rlu - plate_blank_rlu) /
                          (plate_vehicle_interior - plate_blank_rlu)) %>%
  ungroup() %>%
  select(design, plate_id, donor_id, donor_sex, donor_age, compound, weekday,
         well, row, column, is_edge, well_type, conc_uM, replicate,
         raw_rlu, plate_vehicle_interior, plate_blank_rlu, viability_pct,
         ic50_true, failed_plate, tech_cv)

write_csv(ctg, file.path(params$outdir, "ctg_phh_weekday_synthetic.csv"))
glimpse(ctg)
```

## The day effect is visible in the vehicle wells alone

This is the diagnostic you would actually run — it needs no compound data, and
it works identically in both designs.

```{r vehicle-diagnostic, fig.width=7, fig.height=4}
vehicle_summary <- ctg %>%
  filter(well_type == "vehicle", !is_edge, design == DESIGNS[1]) %>%
  group_by(weekday) %>%
  summarise(n = n(), median_rlu = median(raw_rlu), .groups = "drop")

vehicle_summary %>%
  mutate(relative = percentage(median_rlu / min(median_rlu) - 1, fmt = "+1.1"))

p_vehicle <- ctg %>%
  filter(well_type == "vehicle", !is_edge, design == DESIGNS[1]) %>%
  ggplot(aes(weekday, raw_rlu / 1e6, fill = weekday)) +
  geom_boxplot(width = 0.5, outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.12, size = 1.6, alpha = 0.7) +
  scale_fill_manual(values = day_cols, guide = "none") +
  labs(title = "Vehicle wells by run day",
       subtitle = "the reference the whole plate is normalised against",
       x = NULL, y = expression("Raw signal (10"^6~"RLU)")) +
  theme_ctg

p_vehicle
```

# Plate quality control

The QC criteria below are **pre-specified and endpoint-blind**: they use only
the control wells, never the compound wells. That is what makes an exclusion
defensible later.

| Metric | Threshold |
|---|---|
| Z' factor (vehicle vs digitonin) | ≥ 0.50 |
| Vehicle CV (interior wells) | ≤ 15% |
| Signal window (vehicle / blank) | ≥ 10× |
| Vehicle signal vs study median | within 0.5–2× |

```{r plate-qc}
zprime <- function(neg, pos) {
  1 - 3 * (sd(neg) + sd(pos)) / abs(mean(neg) - mean(pos))
}

plate_qc <- ctg %>%
  group_by(design, plate_id, donor_id, compound, weekday) %>%
  summarise(
    vehicle_median = median(raw_rlu[well_type == "vehicle" & !is_edge]),
    vehicle_cv     = sd(raw_rlu[well_type == "vehicle" & !is_edge]) /
                     mean(raw_rlu[well_type == "vehicle" & !is_edge]),
    blank_median   = median(raw_rlu[well_type == "no-cell blank"]),
    window         = vehicle_median / blank_median,
    z_prime        = zprime(raw_rlu[well_type == "vehicle" & !is_edge],
                            raw_rlu[well_type == "lysis control"]),
    .groups = "drop") %>%
  group_by(design) %>%
  mutate(vehicle_rel = vehicle_median / median(vehicle_median)) %>%
  ungroup() %>%
  mutate(qc_pass = z_prime >= 0.5 & vehicle_cv <= 0.15 & window >= 10 &
                   between(vehicle_rel, 0.5, 2))

plate_qc %>%
  filter(design == DESIGNS[1]) %>%
  transmute(plate_id, donor_id, compound, weekday,
            vehicle_rel = round(vehicle_rel, 2),
            vehicle_cv  = percentage(vehicle_cv),
            window      = round(window, 0),
            z_prime     = round(z_prime, 2),
            qc_pass) %>%
  arrange(qc_pass, plate_id)
```

```{r qc-figure, fig.width=7, fig.height=6.4}
p_qc <- plate_qc %>%
  filter(design == DESIGNS[1]) %>%
  pivot_longer(c(z_prime, vehicle_cv, vehicle_rel),
               names_to = "metric", values_to = "value") %>%
  mutate(metric = recode(metric,
                         z_prime     = "Z' factor",
                         vehicle_cv  = "Vehicle CV",
                         vehicle_rel = "Vehicle / median")) %>%
  ggplot(aes(x = fct_reorder(plate_id, value), y = value, fill = qc_pass)) +
  geom_col() +
  geom_hline(data = tibble(metric = c("Z' factor", "Vehicle CV",
                                      "Vehicle / median"),
                           yint = c(0.5, 0.15, 0.5)),
             aes(yintercept = yint), linetype = "dashed", colour = "grey30") +
  facet_wrap(~ metric, scales = "free_x") +
  coord_flip() +
  scale_fill_manual(values = c(`TRUE` = "#3b7a57", `FALSE` = "#a33"),
                    labels = c(`TRUE` = "QC pass", `FALSE` = "QC fail")) +
  labs(title = "Pre-specified plate QC (blocked design)",
       subtitle = "control wells only - the compound data are never inspected at this stage",
       x = NULL, y = NULL) +
  theme_ctg

p_qc
```

One plate fails, and it is the one that was simulated to fail:

```{r qc-caught}
plate_qc %>% filter(!qc_pass) %>% select(design, plate_id, donor_id, compound,
                                         z_prime, vehicle_cv, vehicle_rel)
```

Note what the failure does *not* do. Because the whole plate is scaled down,
normalising to that plate's own vehicle removes the gain almost completely —
the fitted IC~50~ is not strongly biased, it is simply imprecise. QC catches it
through **variance**, not through the mean. A partial failure (one column of
tips blocked) would be the opposite case: it biases the curve shape while
leaving the control wells intact, and is caught by curve-shape checks
(monotonicity, upper asymptote) rather than by Z'.

# Curve fitting

A four-parameter logistic on log~10~ concentration per plate. Two IC~50~
definitions are extracted, and the difference between them matters here:

* **Absolute IC~50~** — the concentration giving 50% of the *vehicle* signal.
  This is the usual reported value and it inherits any shift in the vehicle
  reference.
* **Relative IC~50~** — the midpoint between the fitted upper and lower
  asymptotes, i.e. `xmid`. A uniform scaling of the whole curve is absorbed by
  the upper asymptote, so this estimate is immune to it.

```{r fit}
fit_fpl <- function(df) {
  d <- df %>%
    group_by(conc_uM) %>%
    summarise(viability = mean(viability_pct), .groups = "drop") %>%
    mutate(logc = log10(conc_uM))
  nls(viability ~ SSfpl(logc, A, B, xmid, scal), data = d)
}
safe_fit <- possibly(fit_fpl, otherwise = NULL)

abs_ic50 <- function(fit) {
  if (is.null(fit)) return(NA_real_)
  cf <- coef(fit)
  A <- cf[["A"]]; B <- cf[["B"]]; xmid <- cf[["xmid"]]; scal <- cf[["scal"]]
  if (!(A > 50 && B < 50)) return(NA_real_)
  10^(xmid - scal * log((B - 50) / (50 - A)))
}
rel_ic50 <- function(fit) {
  if (is.null(fit)) return(NA_real_)
  10^(coef(fit)[["xmid"]])
}
top_asym <- function(fit) if (is.null(fit)) NA_real_ else coef(fit)[["A"]]

xmid_se <- function(fit) {
  if (is.null(fit)) return(NA_real_)
  summary(fit)$coefficients["xmid", "Std. Error"]
}

fits <- ctg %>%
  filter(well_type == "test article") %>%
  nest(data = -c(design, plate_id, donor_id, compound, weekday)) %>%
  mutate(fit          = map(data, safe_fit),
         ic50_abs     = map_dbl(fit, abs_ic50),
         ic50_rel     = map_dbl(fit, rel_ic50),
         top_pct      = map_dbl(fit, top_asym),
         se_log_ic50  = map_dbl(fit, xmid_se)) %>%
  left_join(plate_qc %>% select(design, plate_id, qc_pass, z_prime, vehicle_cv),
            by = c("design", "plate_id"))

ic50_tbl <- fits %>%
  select(design, plate_id, donor_id, compound, weekday, ic50_abs, ic50_rel,
         top_pct, se_log_ic50, qc_pass) %>%
  left_join(truth %>% select(donor_id, compound, ic50_true),
            by = c("donor_id", "compound")) %>%
  arrange(design, compound, donor_id)

write_csv(ic50_tbl, file.path(params$outdir, "ctg_phh_weekday_ic50.csv"))
ic50_tbl %>% mutate(across(where(is.numeric), ~ round(.x, 2))) %>% head(10)
```

The fitted upper asymptote already tells the story: on Friday plates the curve
tops out near
`r round(100 / (1 + params$friday_vehicle_gain))`%, not 100%.

```{r top-asymptote}
ic50_tbl %>%
  filter(qc_pass) %>%
  group_by(weekday) %>%
  summarise(mean_top_pct = round(mean(top_pct, na.rm = TRUE), 1),
            n = n(), .groups = "drop")
```

# Consequences of the two designs

The day-effect analysis below uses **QC-passing plates only**; the failed plate
is dealt with in its own section.

```{r gm-helpers}
ic50_qc <- ic50_tbl %>% filter(qc_pass)

gm  <- function(x) exp(mean(log(x), na.rm = TRUE))
gsd <- function(x) exp(sd(log(x), na.rm = TRUE))
true_sep <- params$ic50_B / params$ic50_A
```

## Recovered potency

```{r recovery-table}
recovery <- ic50_qc %>%
  group_by(design, compound) %>%
  summarise(GM_absolute = gm(ic50_abs),
            GM_relative = gm(ic50_rel),
            GM_true     = gm(ic50_true), .groups = "drop") %>%
  mutate(bias_absolute = percentage(GM_absolute / GM_true - 1, fmt = "+1.1"),
         bias_relative = percentage(GM_relative / GM_true - 1, fmt = "+1.1"),
         across(starts_with("GM"), ~ round(.x, 1)))

recovery
```

## Apparent separation between the compounds

The quantity the study exists to estimate.

```{r separation}
separation <- ic50_qc %>%
  select(design, donor_id, compound, ic50_abs, ic50_rel) %>%
  pivot_wider(names_from = compound, values_from = c(ic50_abs, ic50_rel)) %>%
  mutate(sep_absolute = `ic50_abs_CPD-B` / `ic50_abs_CPD-A`,
         sep_relative = `ic50_rel_CPD-B` / `ic50_rel_CPD-A`)

sep_summary <- separation %>%
  group_by(design) %>%
  summarise(GM_separation_absolute = gm(sep_absolute),
            GM_separation_relative = gm(sep_relative),
            .groups = "drop") %>%
  mutate(true_separation = true_sep,
         inflation = percentage(GM_separation_absolute / true_sep - 1,
                                fmt = "+1.1"),
         across(where(is.numeric), ~ round(.x, 2)))

sep_summary
```

## Is the day effect estimable at all?

The decisive difference. In the blocked design, compound and weekday are
separate columns of the model matrix and both coefficients are estimated. In
the confounded design they are perfectly aliased, and `lm` returns `NA` for the
second term — R is telling you the design cannot answer the question.

```{r aliasing}
model_tbl <- ic50_qc %>%
  mutate(log_ic50 = log10(ic50_abs))

fit_blocked <- lm(log_ic50 ~ compound + weekday,
                  data = filter(model_tbl, design == DESIGNS[1]))
fit_confounded <- lm(log_ic50 ~ compound + weekday,
                     data = filter(model_tbl, design == DESIGNS[2]))

bind_rows(
  tidy(fit_blocked)    %>% mutate(design = DESIGNS[1]),
  tidy(fit_confounded) %>% mutate(design = DESIGNS[2])
) %>%
  select(design, term, estimate, std.error, p.value) %>%
  mutate(across(where(is.numeric), ~ signif(.x, 3)))

## the aliasing, stated explicitly
alias(fit_confounded)
```

# Figures

```{r results-figure, fig.width=7, fig.height=17.5}
cpd_cols <- c("CPD-A" = "#1f6fb4", "CPD-B" = "#c1443b")

grid <- tibble(logc = seq(log10(0.2), log10(1500), length.out = 200))
curves <- fits %>%
  filter(!map_lgl(fit, is.null)) %>%
  mutate(pred = map(fit, ~ mutate(grid, viability = predict(.x, newdata = grid)))) %>%
  select(design, donor_id, compound, weekday, pred) %>%
  unnest(pred) %>%
  mutate(conc_uM = 10^logc)

## A: curves, faceted by design
pA <- ggplot(curves, aes(conc_uM, viability, colour = compound)) +
  geom_line(aes(group = interaction(donor_id, compound)), alpha = 0.7,
            linewidth = 0.8) +
  facet_wrap(~ design) +
  geom_hline(yintercept = 50, linetype = "dashed", colour = "grey40") +
  geom_hline(yintercept = 100, linetype = "dotted", colour = "grey60") +
  scale_x_log10(labels = scales::label_number(drop0trailing = TRUE)) +
  scale_colour_manual(values = cpd_cols) +
  coord_cartesian(ylim = c(-5, 118)) +
  labs(title = "A  Concentration–response by design",
       subtitle = "confounded: every CPD-A curve tops out below 100%",
       x = "Concentration (µM)", y = "ATP (% of vehicle)") +
  theme_ctg + theme(legend.position = "bottom")

## B: absolute IC50 vs truth
pB <- ic50_qc %>%
  ggplot(aes(ic50_true, ic50_abs, colour = compound, shape = weekday)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(size = 2, alpha = 0.9) +
  facet_wrap(~ design) +
  scale_x_log10() + scale_y_log10() +
  scale_colour_manual(values = cpd_cols) +
  labs(title = "B  Recovered vs simulated IC50 (absolute)",
       subtitle = "below the diagonal = biased toward apparent toxicity",
       x = expression("Simulated IC"[50]~"(µM)"),
       y = expression("Fitted IC"[50]~"(µM)")) +
  theme_ctg + theme(legend.position = "bottom")

## C: per-donor separation
pC <- separation %>%
  ggplot(aes(design, sep_absolute)) +
  geom_hline(yintercept = true_sep, linetype = "dashed", colour = "grey30",
             linewidth = 0.9) +
  geom_line(aes(group = donor_id), colour = "grey75", linewidth = 0.6) +
  geom_point(aes(colour = design), size = 2.4, show.legend = FALSE) +
  stat_summary(fun = gm, fun.min = gm, fun.max = gm, geom = "crossbar",
               width = 0.4, linewidth = 0.8, colour = "grey20") +
  annotate("text", x = 0.6, y = true_sep * 1.06,
           label = sprintf("true separation %.0f×", true_sep),
           hjust = 0, size = GEOM_PT, colour = "grey25") +
  scale_colour_manual(values = setNames(c("#3b7a57", "#a33"), DESIGNS)) +
  scale_y_log10() +
  labs(title = "C  Apparent CPD-B / CPD-A separation, per donor",
       x = NULL, y = "Fold separation (absolute IC50)") +
  theme_ctg

## D: absolute vs relative IC50 bias
pD <- ic50_qc %>%
  pivot_longer(c(ic50_abs, ic50_rel), names_to = "metric", values_to = "ic50") %>%
  mutate(metric = recode(metric, ic50_abs = "absolute IC50",
                         ic50_rel = "relative IC50"),
         rel_bias = ic50 / ic50_true - 1) %>%
  ggplot(aes(metric, rel_bias, fill = weekday)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_boxplot(width = 0.55, alpha = 0.8, outlier.shape = NA) +
  facet_wrap(~ design) +
  scale_fill_manual(values = day_cols) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "D  Bias of the two IC50 definitions",
       subtitle = "the relative IC50 absorbs the shift into the upper asymptote",
       x = NULL, y = "Relative bias vs simulated truth") +
  theme_ctg + theme(legend.position = "bottom")

fig <- pA / pB / pC / pD
fig
```

```{r export-figures}
openFileDevice(file.path(params$outdir, "ctg_weekday_results.pdf"),
               width = 7, height = 17.5)
print(fig)
closeFileDevice()

openFileDevice(file.path(params$outdir, "ctg_weekday_design.pdf"),
               width = 7, height = 4)
print(p_design)
closeFileDevice()

openFileDevice(file.path(params$outdir, "ctg_plate_layout.pdf"),
               width = 7, height = 3.6)
print(p_layout)
closeFileDevice()
```

# Handling the three perturbations

Everything below uses the **blocked** design, so the run-day issue is out of
the way and the three perturbations can be looked at on their own.

```{r blocked-subset}
blocked <- ic50_tbl %>% filter(design == DESIGNS[1])
failed_donors <- blocked %>% filter(!qc_pass) %>% pull(donor_id) %>% as.character()
failed_donors
```

## Perturbation 1 — the technically failed plate

Three defensible ways to handle it, and one indefensible one (dropping it
because the number looked wrong).

```{r failed-plate-scenarios}
scenarios <- bind_rows(
  blocked %>% mutate(scenario = "1. keep everything"),
  blocked %>% filter(qc_pass) %>% mutate(scenario = "2. drop the failed plate"),
  blocked %>% filter(!as.character(donor_id) %in% failed_donors) %>%
    mutate(scenario = "3. drop the whole donor")
) %>%
  mutate(scenario = ofactor(scenario))

scenario_summary <- scenarios %>%
  group_by(scenario, compound) %>%
  summarise(n_plates = sum(!is.na(ic50_rel)),
            GM_IC50  = gm(ic50_rel),
            GSD      = gsd(ic50_rel), .groups = "drop") %>%
  mutate(across(c(GM_IC50, GSD), ~ round(.x, 2)))

scenario_summary
```

The paired contrast is where the difference bites: dropping a single plate
leaves that donor with one compound only, so it contributes to the
compound-level means but not to the pairing.

```{r failed-plate-separation}
sep_scenarios <- scenarios %>%
  select(scenario, donor_id, compound, ic50_rel) %>%
  pivot_wider(names_from = compound, values_from = ic50_rel) %>%
  mutate(ratio = `CPD-B` / `CPD-A`) %>%
  group_by(scenario) %>%
  summarise(n_donors_with_both = sum(!is.na(ratio)),
            GM_separation = gm(ratio),
            GSD_separation = gsd(ratio),
            .groups = "drop") %>%
  mutate(true_separation = true_sep,
         across(where(is.numeric), ~ round(.x, 2)))

sep_scenarios
```

Keeping the failed plate does not shift the point estimate much — the
whole-plate gain cancels in the normalisation — but it inflates the spread,
because that plate's IC~50~ carries a standard error several times larger than
the others. Dropping the plate is the right call, and it is the *QC rule*, not
the IC~50~, that justifies it.

## Perturbation 2 — the sensitive donor

Donor `r params$sensitive_donor` is `r scales::percent(1 - params$sensitive_factor)`
more sensitive to both compounds. A routine outlier test will happily flag it.

```{r grubbs}
grubbs <- function(x, alpha = 0.05) {
  x <- x[!is.na(x)]; n <- length(x)
  G <- max(abs(x - mean(x))) / sd(x)
  tcrit <- qt(alpha / (2 * n), df = n - 2)
  Gcrit <- (n - 1) / sqrt(n) * sqrt(tcrit^2 / (n - 2 + tcrit^2))
  tibble(n = n, G = G, G_crit = Gcrit, flagged = G > Gcrit,
         extreme = names(which.max(abs(x - mean(x)))))
}

blocked %>%
  filter(qc_pass) %>%
  group_by(compound) %>%
  group_modify(~ grubbs(setNames(log10(.x$ic50_rel), as.character(.x$donor_id)))) %>%
  ungroup() %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

What removing it would do:

```{r sensitive-donor-effect}
with_without <- bind_rows(
  blocked %>% filter(qc_pass) %>% mutate(set = "all donors"),
  blocked %>% filter(qc_pass,
                     as.character(donor_id) != params$sensitive_donor) %>%
    mutate(set = "sensitive donor removed")
) %>%
  group_by(set, compound) %>%
  summarise(n = n(), GM_IC50 = gm(ic50_rel), GSD = gsd(ic50_rel),
            most_sensitive_donor_IC50 = min(ic50_rel, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))

with_without
```

And what it does to the quantity the study is actually estimating:

```{r sensitive-donor-ratio}
blocked %>%
  filter(qc_pass) %>%
  select(donor_id, compound, ic50_rel) %>%
  pivot_wider(names_from = compound, values_from = ic50_rel) %>%
  mutate(ratio = `CPD-B` / `CPD-A`,
         sensitive = as.character(donor_id) == params$sensitive_donor) %>%
  arrange(desc(sensitive)) %>%
  mutate(across(where(is.numeric), ~ round(.x, 2)))
```

Because the donor is shifted on *both* compounds, the within-donor ratio is
untouched. The pairing absorbs the donor entirely — which is the strongest
practical argument for testing both compounds in every donor.

## Perturbation 3 — unequal technical variance

Replicate spread is not constant: it differs between hepatocyte lots and it
grows as the signal falls towards the assay floor.

```{r variance-structure}
rep_stats <- ctg %>%
  filter(design == DESIGNS[1], well_type == "test article") %>%
  group_by(donor_id, compound, conc_uM) %>%
  summarise(mean_viab = mean(viability_pct),
            sd_viab   = sd(viability_pct),
            n = n(), .groups = "drop") %>%
  left_join(donor_noise, by = "donor_id") %>%
  mutate(lot = if_else(tech_cv > 0.08, "high-variance lot", "standard lot"))

rep_stats %>%
  group_by(lot) %>%
  summarise(median_sd = round(median(sd_viab, na.rm = TRUE), 2),
            n_conditions = n(), .groups = "drop")
```

### Weighting the fit instead of discarding data

```{r weighted-fit}
fit_fpl_weighted <- function(df) {
  d <- df %>%
    group_by(conc_uM) %>%
    summarise(viability = mean(viability_pct),
              s = sd(viability_pct), n = n(), .groups = "drop") %>%
    mutate(logc = log10(conc_uM), w = n / pmax(s, 1)^2)
  nls(viability ~ SSfpl(logc, A, B, xmid, scal), data = d, weights = w)
}
safe_fit_w <- possibly(fit_fpl_weighted, otherwise = NULL)

weighted_tbl <- ctg %>%
  filter(design == DESIGNS[1], well_type == "test article") %>%
  nest(data = -c(plate_id, donor_id, compound)) %>%
  mutate(fit_w      = map(data, safe_fit_w),
         ic50_w     = map_dbl(fit_w, rel_ic50),
         se_w       = map_dbl(fit_w, xmid_se)) %>%
  select(plate_id, donor_id, compound, ic50_w, se_w) %>%
  left_join(blocked %>% select(plate_id, ic50_rel, se_log_ic50, qc_pass,
                               ic50_true),
            by = "plate_id") %>%
  filter(qc_pass)

weighted_tbl %>%
  group_by(compound) %>%
  summarise(median_se_unweighted = round(median(se_log_ic50, na.rm = TRUE), 4),
            median_se_weighted   = round(median(se_w, na.rm = TRUE), 4),
            median_abs_error_unweighted =
              round(median(abs(log10(ic50_rel / ic50_true)), na.rm = TRUE), 4),
            median_abs_error_weighted =
              round(median(abs(log10(ic50_w / ic50_true)), na.rm = TRUE), 4),
            .groups = "drop")
```

### Separating biological from technical variability

The across-donor GSD everyone quotes is a *sum* of real donor differences and
assay noise. With a per-plate standard error in hand, the two can be separated
by moments: `Var(log IC50) = Var_biological + mean(SE²)`.

```{r variance-decomposition}
var_decomp <- blocked %>%
  filter(qc_pass) %>%
  group_by(compound) %>%
  summarise(var_total = var(log10(ic50_rel), na.rm = TRUE),
            var_technical = mean(se_log_ic50^2, na.rm = TRUE),
            .groups = "drop") %>%
  mutate(var_biological = pmax(var_total - var_technical, 0),
         GSD_observed   = 10^sqrt(var_total),
         GSD_biological = 10^sqrt(var_biological),
         technical_share = percentage(var_technical / var_total))

var_decomp %>% mutate(across(where(is.numeric), ~ round(.x, 3)))
```

### A variance-weighted estimate of the separation

```{r weighted-separation}
pairs_tbl <- blocked %>%
  filter(qc_pass) %>%
  select(donor_id, compound, ic50_rel, se_log_ic50) %>%
  pivot_wider(names_from = compound,
              values_from = c(ic50_rel, se_log_ic50)) %>%
  filter(!is.na(`ic50_rel_CPD-A`), !is.na(`ic50_rel_CPD-B`)) %>%
  mutate(d = log10(`ic50_rel_CPD-B`) - log10(`ic50_rel_CPD-A`),
         v = `se_log_ic50_CPD-A`^2 + `se_log_ic50_CPD-B`^2,
         w = 1 / v)

unw <- mean(pairs_tbl$d)
wtd <- weighted.mean(pairs_tbl$d, pairs_tbl$w)

tibble(estimator = c("unweighted", "inverse-variance weighted"),
       log10_separation = c(unw, wtd),
       fold_separation = 10^c(unw, wtd),
       true_fold = true_sep) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3)))
```

# Figures for the perturbations

```{r perturbation-figure, fig.width=7, fig.height=17.5}
## E: variance is not constant
pE <- rep_stats %>%
  ggplot(aes(mean_viab, sd_viab, colour = lot)) +
  geom_point(size = 1.7, alpha = 0.75) +
  geom_smooth(method = "loess", se = FALSE, linewidth = 1.1, span = 1) +
  scale_colour_manual(values = c("high-variance lot" = "#c1443b",
                                 "standard lot" = "#1f6fb4")) +
  labs(title = "E  Replicate spread depends on lot and on signal level",
       subtitle = "equal-weight fitting treats all points as equally informative",
       x = "Mean viability of the condition (% of vehicle)",
       y = "SD across replicate wells") +
  theme_ctg + theme(legend.position = "bottom")

## F: what each exclusion scenario does to the paired separation
pF <- scenarios %>%
  select(scenario, donor_id, compound, ic50_rel) %>%
  pivot_wider(names_from = compound, values_from = ic50_rel) %>%
  mutate(ratio = `CPD-B` / `CPD-A`) %>%
  filter(!is.na(ratio)) %>%
  ggplot(aes(scenario, ratio)) +
  geom_hline(yintercept = true_sep, linetype = "dashed", colour = "grey30",
             linewidth = 0.9) +
  geom_point(size = 2.2, alpha = 0.8, colour = "#3b7a57") +
  stat_summary(fun = gm, fun.min = gm, fun.max = gm, geom = "crossbar",
               width = 0.4, colour = "grey20") +
  scale_y_log10() +
  labs(title = "F  Effect of each exclusion rule on the paired separation",
       subtitle = sprintf("dashed line = true %.0f-fold separation", true_sep),
       x = NULL, y = "CPD-B / CPD-A (per donor)") +
  theme_ctg +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

## G: the sensitive donor in context
pG <- blocked %>%
  filter(qc_pass) %>%
  mutate(group = case_when(
    as.character(donor_id) == params$sensitive_donor ~ "sensitive donor",
    TRUE ~ "other donors")) %>%
  ggplot(aes(x = fct_reorder(donor_id, ic50_rel), y = ic50_rel,
             colour = compound, shape = group)) +
  geom_point(size = 2.4) +
  scale_shape_manual(values = c("sensitive donor" = 17, "other donors" = 16)) +
  scale_colour_manual(values = cpd_cols) +
  scale_y_log10() +
  labs(title = "G  Donor sensitivity is real, and it is correlated across compounds",
       subtitle = "low for both compounds, so the ratio is unaffected",
       x = "Donor", y = expression("Relative IC"[50]~"(µM)")) +
  theme_ctg +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")

## H: variance decomposition
pH <- var_decomp %>%
  select(compound, var_biological, var_technical) %>%
  pivot_longer(-compound, names_to = "source", values_to = "variance") %>%
  mutate(source = recode(source, var_biological = "biological",
                         var_technical = "technical")) %>%
  ggplot(aes(compound, variance, fill = source)) +
  geom_col(width = 0.6) +
  scale_fill_manual(values = c(biological = "#3b7a57", technical = "#d9a441")) +
  labs(title = "H  What the across-donor spread is made of",
       subtitle = "the quoted GSD mixes donor biology with assay noise",
       x = NULL, y = expression("Variance of log"[10]~"IC"[50])) +
  theme_ctg + theme(legend.position = "bottom")

fig2 <- pE / pF / pG / pH
fig2
```

```{r export-perturbation-figure}
openFileDevice(file.path(params$outdir, "ctg_perturbations.pdf"),
               width = 7, height = 17.5)
print(fig2)
closeFileDevice()

openFileDevice(file.path(params$outdir, "ctg_plate_qc.pdf"),
               width = 7, height = 6.4)
print(p_qc)
closeFileDevice()
```

# Remove, keep, or model?

The three perturbations look superficially similar — each produces values that
sit away from the rest — and they call for three different responses.

**The failed plate: remove it, on a rule written beforehand.** The exclusion is
justified by control wells (Z', vehicle CV, signal window), not by the IC~50~.
That ordering is the whole point: a rule applied before the endpoint is seen
cannot be tuned, consciously or otherwise, to produce a preferred answer. Write
the thresholds into the protocol, report how many plates failed and why, and
keep the raw data in the deposited dataset with the flag attached rather than
silently deleting the rows. Where the failed plate belongs to a paired design,
prefer dropping the plate and using an estimator that tolerates the imbalance
over dropping the whole donor — throwing away a perfectly good CPD-A plate to
preserve symmetry costs information for no gain, unless the analysis is a
strictly paired test.

**The sensitive donor: keep it.** Grubbs flags it, and Grubbs is answering the
wrong question. The test asks whether the value is improbable under an assumed
distribution; it cannot tell you whether the value is *wrong*. Nothing in the
QC record marks that plate as abnormal, the donor is shifted coherently on both
compounds, and a lot that is genuinely more sensitive is exactly the
observation a safety margin should be built on. Removing it shrinks the
reported GSD, which looks like a cleaner result while actually deleting the
information that mattered. If a donor must be set aside, the reason has to be
technical and documented — not "it was an outlier".

**Unequal variance: model it, do not prune it.** Heteroscedasticity is not
contamination; it is a property of the assay. The noisy lots still carry
information, just less of it per well, and the fix is to weight accordingly:
weighted nls within a plate, inverse-variance weighting across donors. The more
consequential habit is to stop quoting a bare across-donor GSD. Once the
per-plate standard errors are available, splitting that spread into biological
and technical components takes one line and changes how the number should be
read — a GSD that is largely technical is an argument for more replicates, not
for a wider safety factor.

**A rule of thumb that covers all three.** Decide the question first: is the
value *implausible given the controls*, or merely *inconvenient given the
hypothesis*? The first is a QC exclusion; the second is data manipulation. In
practice, running the analysis both ways and reporting both is cheap — every
scenario table above is three lines of `dplyr` — and a result that survives
both is worth considerably more than one that requires a particular exclusion.

# What the design simulation shows

* **The bias is predictable in size.** A vehicle reference inflated by a factor
  *g* scales every normalised viability on that plate by 1/*g*. For a 4PL with
  Hill slope *h*, the absolute IC~50~ moves by a factor
  ((2/*g*) − 1)^1/h^ — about
  `r round(((2 / (1 + params$friday_vehicle_gain)) - 1)^(1 / params$hill_A), 3)`
  for CPD-A and
  `r round(((2 / (1 + params$friday_vehicle_gain)) - 1)^(1 / params$hill_B), 3)`
  for CPD-B at the simulated `r scales::percent(params$friday_vehicle_gain)`
  gain. Nothing about the compounds changed.
* **Confounding turns that bias into a conclusion.** With CPD-A on Friday and
  CPD-B on Monday, only CPD-A is deflated, so the apparent separation is
  inflated above the true `r round(true_sep, 1)`-fold. The result looks like
  sharper discrimination between the compounds, which is exactly the kind of
  error that survives review.
* **Blocking does not remove the bias — it removes the confounding.** In the
  blocked design both compounds are equally affected on Friday plates, so the
  contrast is unbiased even though individual absolute IC~50~ values on those
  plates still are. Day becomes an estimable nuisance term rather than an
  invisible one.
* **`alias()` is the check worth running.** Before the assay, not after: if a
  term drops out of the model matrix, no amount of analysis recovers it.
* **The upper asymptote is a free diagnostic.** A fitted top well below 100%
  when the compound has a true no-effect plateau points at the vehicle
  reference, not at the compound. Reporting the relative IC~50~ alongside the
  absolute one makes the discrepancy visible.
* **Stronger still: block within the plate.** Putting both compounds on the same
  plate, as in the shared-plate layout, makes the day effect cancel exactly in
  the compound contrast, because both arms then share one vehicle reference.
  The separate-plate layout used here is the weakest arrangement that can still
  be salvaged by scheduling.

# Session information

```{r session}
sessionInfo()
```
